Replication stress inducing ELF3 upregulation promotes BRCA1-deficient breast tumorigenesis in luminal progenitors

Version 1

Abstract

Abstract

BRCA1 is a critical tumor suppressor, mutations in which greatly increase risks for many tumors in carriers, most notably breast cancer. Luminal progenitor cells (LPs) are the currently recognized cells origin of BRCA1-deficient breast cancers. However, the reason why LPs are prone to transform with BRCA1 deficiency has not been elucidated. Here, using single-cell sequencing of human BRCA1 mutant breast cancers and RNA sequencing (RNA-seq) of BRCA1-deficient normal mammary cells, we reveal that replication stress is a feature of LPs and a driving factor during BRCA1-associated tumorigenesis. Mechanistically, replication stress and BRCA1 deficiency lead to significant upregulation of ELF3 expression. ELF3 can help suppress excessive genomic instability and promote LP transformation with BRCA1 deficiency. Moreover, ELF3 emerged as a core transcription factor regulating LP genes, leading to LP expansion. Our findings suggest that replication stress is a driving factor during BRCA1-associated tumorigenesis in luminal progenitor cells and elucidates the key role of ELF3 during this process.

Introduction

BRCA1 is one of the most important tumor suppressor genes. Carriers of BRCA1 germline mutations are at greatly increased risks for developing various tumors, including breast, ovarian, prostate, colon, pancreatic, and gastric cancers (1, 2).Notably, BRCA1 is most closely associated with breast cancer. However, why BRCA1-associated tumors exhibit significant tissue specificity is unknown, and the effects of BRCA1 defects in normal human mammary cells are largely obscure.

In addition to the tissue specificity of tumors caused by BRCA1 deficiency, breast cancers associated with BRCA1 deficiency exhibit a set of other distinctive features. Most BRCA1 mutation carriers develop triple-negative breast cancers (TNBCs), in which tumors show negative expression of estrogen receptor (ER), progesterone receptor (PR), and Her2 (human epidermal growth factor receptor 2), and these cancers are usually also classified as basal-like breast cancers. The currently recognized cells of origin of BRCA1-deficient breast cancers are the luminal progenitor cell (LPs) in the mammary gland (3-6). LPs show abnormal expansion in the normal mammary tissue of BRCA1 mutation carriers (3, 7), which might be caused by a differentiation blockade (8). Investigating the traits of LPs is important for tumor prevention and treatment in BRCA1 mutation carriers and a wider range of patients with TNBCs and basal-like breast cancer. However, the reason LPs are prone to transform into cancer cells when BRCA1 is deficient is unclear.

ELF3 is a member of the ETS family of transcription factors. Early reports suggested that ELF3 may regulate the development and differentiation of mammary tissue (9). Intriguingly, ELF3 has opposite functions in tumorigenesis in different tissues. In breast cancers, ELF3 shows high expression (10), plays an oncogenic role by promoting EMT (11) and forms a positive feedback loop with Her2 to maintain the transformation phenotype of Her2+ breast cancers (12). However, the mechanism of high ELF3 expression in certain types of breast cancer and its association with breast cancers beyond the Her2+ subtype is currently unknown.

In this study, utilizing single-cell sequencing of BRCA1 mutant breast cancers and time-resolved RNA-seq of BRCA1-deficient normal mammary cells, we revealed that LPs have a trait of higher replication stress compared to other cell populations in the normal mammary tissue, endowing them with the potential for transformation when BRCA1 is deficient, and ELF3 plays a key role in the process. Mechanistically,

ELF3 expression is significantly upregulated by replication stress and BRCA1 deficiency, dependent on the ATR-Chk1-E2F axis and the transcriptional regulation function of BRCA1, respectively. In addition, ELF3 can help suppress excessive genomic instability, promoting the tolerance of LPs to BRCA1 deficiency. Notably, BRCA1 deficiency causes cells to display LP transcriptional profile characteristics, leading to dedifferentiation and expansion of LPs, and ELF3 is a core transcription factor during this process. Our study reveals why BRCA1 deficiency is prone to result in tumorigenesis in LPs and elucidates the key role of replication stress and ELF3 during this process, suggesting that ELF3 could be a promising target for BRCA1-associated breast cancers.

Results

LPs have higher levels of replication stress compared with other mammary cell populations

LPs in the mammary gland are the currently recognized cells of origin of BRCA1-deficient breast cancers. However, the reason for LPs are prone to transform with BRCA1 deficiency has not been elucidated. To address this question, we attempt to investigate from two aspects that may influence the process of tumorigenesis. Firstly, we considered that LPs, which are stimulated by upstream progesterone paracrine signals through RANKL (5), may have higher levels of replication stress, which can trigger tumorigenesis by increasing genomic instability. We utilized both published data and our published single-cell sequencing data (8) to address this question. We analyzed the data of Lim et al. (3) by scoring multiple DNA replication pathway gene expression to infer the replication stress in different mammary cell populations as previously studied (13). The result showed that LPs have the highest level of replication stress among all subgroups of normal human mammary cells (Fig. 1A). Our single-cell sequencing data were obtained from the tumor cells and normal mammary cells of four BRCA1 mutation carriers (case #1-4) and normal mammary cells of three matched non-carriers (case #5-7). According to our previous analysis, case #3 BRCA1 mutant carrier developed basal-like breast cancer, which was derived from LPs. Importantly, in this case of BRCA1 mutant carrier, single-cell sequencing data revealed that tumor cells had the highest levels of replication stress, and LPs had the second-highest levels (Fig. 1B). Moreover, we applied Monocle to reconstruct the transformation trajectories from normal mammary cells to tumor cells in the BRCA1 mutation carrier and found that the levels of replication stress showed an increasing trend during the process of LP transformation into tumor cells (Fig. 1C and Supplemental Fig. S1A). These data demonstrate that replication stress is an important trait of LPs and might be a key driver of BRCA1-deficient tumorigenesis, which is consistent with the previous report that BRCA1 mutant mammary epithelial cells are defective in replication stress suppression (14).

LPs have higher levels of replication stress compared with other mammary cell populations.

(A) Replication stress pathway scores of different human normal mammary cell populations. The gene expression data were from Lim et al. (3).

(B) T-distributed statistical neighbor embedding (t-SNE) plots visualizing the replication stress scores and ELF3 expression levels in breast cancer and normal mammary tissue from a BRCA1-mutated triple-negative breast cancer patient. The cells are colored by cell type (top) and replication stress score (bottom).

(C) Monocle-generated cell trajectories visualizing the change in replication stress scores during tumorigenesis. The tumor cells and compared normal luminal cells were obtained from a triple-negative breast cancer patient (case #3) carrying a BRCA1 germline mutation. LP, luminal progenitor cells; ML, mature luminal cells.

(D) Volcano plots showing DEGs in MCF10A-shBRCA1-Tet-on cell RNA-seq results after 1 μg/mL DOX treatment for 2 (left), 5 (middle) and 10 (right) days versus the Ctrl group. DEGs are genes with |log2-fold change| > 2 and FDR < 0.001.

(E) PCA plot of MCF10A-shBRCA1-Tet-on cell RNA-seq results.

(F) Mfuzz clustering analysis of Day 10 versus Ctrl DEGs. The heatmap (left) shows the differential expression mode of all genes in each cluster. The line chart (right) shows the global changing mode of each cluster.

(G) Overlap of HU treated MCF10A RNA-seq results and MCF10A-shBRCA1-Tet-on cell RNA-seq Cluster1 genes.

Next, we aimed to find other key regulators during BRCA1-deficient tumorigenesis in LPs. In addition to inducing replication stress, BRCA1 deficiency also profoundly affects the transcriptional profile of normal mammary cells by disrupting the transcriptional function of BRCA1, which may further boost the process of tumorigenesis. To investigate the transcriptional impact of BRCA1 deficiency in normal human mammary epithelial cells, we took advantage of the Tet-on system to build an inducible knockdown model in MCF10A cells. By adjusting the treatment condition of DOX (doxycycline), different time points of BRCA1 deficiency can be controlled. Treatment with DOX causes a constant and efficient knockdown of BRCA1, verified at mRNA and protein levels (Supplemental Fig. S1B). We showed that DOX treatment induces significant growth arrest (Supplemental Fig. S1C) and an increase in genomic instability (Supplemental Fig. S1D), consistent with the phenotypes of BRCA1-deficient cells (15), confirming that our system functioned as expected. We then performed time-resolved RNA-seq with different durations of DOX treatment (2, 5 and 10 days; Supplemental Fig. S1E). Notably, 10 days of DOX treatment resulted in growth arrest, thus cells collected for sequencing have gained growth advantage. Differential gene expression analysis revealed that BRCA1 was efficiently knocked down during the whole experiment (Fig. 1D). Importantly, the principal component analysis (PCA) revealed that samples showed a time-course distribution in the dimension of PC1, which has the dominant power of explanation (Fig. 1E and Supplemental Fig. S1F). These data demonstrate that the variations between samples can be largely attributed to time. To further investigate the modes of differentially expressed genes (DEGs) that change over time, we carried out the DEG cluster analysis using Mfuzz (16). Analysis of the Day 10 DEGs showed that genes could be divided into 5 clusters with different expression change patterns (Fig. 1F). We mainly focused on Cluster 1, for these sharply upregulated genes may play more important roles in cells gaining growth advantage with BRCA1 deficiency at Day 10, mimicking the initiation of BRCA1 deficiency associated tumorigenesis. To find out the most potent regulator upregulated both by replication stress and BRCA1 deficiency, we performed RNA-seq in HU-treated MCF10A cells and overlapped the upregulated genes in HU RNA-seq and Cluster1 genes in BRCA1 deficiency RNA-seq (Fig. 1G). The results showed that ELF3, a transcription factor, is one of the most dominant genes in this analysis. ELF3 has been reported to be involved in functions including EMT in human breast cancer and promotes the transformation of normal human mammary cells (11, 17). Therefore, we speculated that ELF3 could be an important driving factor during the tumorigenesis of BRCA1-deficient and basal-like breast cancer.

ELF3 is upregulated in BRCA1-associated breast cancer and is related to a worse prognosis

We next explored the expression of ELF3 in breast cancers in the TCGA and METABRIC databases to further investigate the association between ELF3 and BRCA1 and validate our previous analysis. In both clinical databases, ELF3 expression is significantly higher in BRCA1-associated breast tumors than in non-BRCA1-associated breast tumors (Fig. 2A). Moreover, the expression levels of BRCA1 and ELF3 showed a significant negative correlation, consistent with our findings (Fig. 2B). Since most BRCA1 mutation carriers develop basal-like breast cancers, we next investigated the association between ELF3 and basal-like breast cancers. We exploited the TCGA and METABRIC databases to compare ELF3 expression levels among different breast cancer subtypes. As we expected, ELF3 showed the highest expression levels in basal-like breast cancer in both databases (Fig. 2C). We took advantage of novel subgroups identified by the METABRIC (18) and further explored ELF3 expression in ten IntClusts of the METABRIC database. Among all subgroups, IntClust 5 was enriched with Her2+ cancers, and IntClust 10 was enriched with basal-like cancers. These two subgroups showed the highest expression levels of ELF3, with no significant difference between them (Fig. 2D). The higher levels of ELF3 expression in Her2 subtype breast cancer are consistent with previous studies showing that ELF3 may be a downstream transcription target of Her2 and that there might be a positive feedback loop between them (12, 19, 20). On the other hand, IntClust 4 shows a favorable outcome and has a ‘CNA-devoid’ feature, which has the lowest level of ELF3 expression. These data are consistent with our findings that ELF3 is upregulated by BRCA1 loss, indicate that ELF3 may play an important role in basal-like breast tumorigenesis, and moreover, the expression of ELF3 may be related to genomic stability status. Furthermore, among human breast cancer tissue samples, ELF3 expression levels were significantly higher in TNBCs than in other subtypes (Fig. 2E and F). Additionally, TNBCs have higher ELF3 expression levels than the Her2 subtype (Fig. 2E and F), revealing a tighter connection between ELF3 and BRCA1-associated breast cancers.

ELF3 is upregulated in BRCA1-associated breast cancers and is related to a worse prognosis.

(A) ELF3 expression levels in BRCA1-associated breast cancers and BRCA1-nonassociated breast cancers in the TCGA and METABRIC databases. BRCA1-associated breast cancers comprise samples with BRCA1 mutations, heterozygous loss and homozygous deletion (Mann–Whitney U test).

(B) Correlation of ELF3 and BRCA1 expression levels in TCGA breast cancer datasets.

(C) ELF3 expression levels of different molecular subtypes of breast cancers in the TCGA and METABRIC databases (one-way ANOVA).

(D) ELF3 expression levels of different IntClusts of breast cancers in the METABRIC database (one-way ANOVA; for IntClust 5 and 10, Mann–Whitney U test, ns, no significance).

(E) Immunohistochemistry images of human triple-negative, Her2+ and ER+ PR+ Her2+ breast cancer samples stained with the ELF3 antibody.

(F) ELF3 expression levels of the indicated types of human breast cancers (Mann– Whitney U test, *, p < 0.05).

(G) Survival analysis of the indicated subtypes of human breast cancer in the KMplot database. For the TNBC group, samples with negative expression of all hormone receptors (ER, PR and Her2) are included. In the TNBC group, high expression (n=54), low expression (n=62); in the basal group, high expression (n=85), low expression (n=211); in the Lum A group, high expression (n=121), low expression (n=101); in the Lum B group, high expression (n=133), low expression (n=67); and in the Her2 group, high expression (n=72), low expression (n=126).

ELF3 shows significant copy number amplification in breast cancer (21). Thus, we wondered that whether the high expression levels of ELF3 in BRCA1-associated breast cancers are due to copy number variation. Therefore, we analyzed the proportion of ELF3 copy number amplification in different subtypes (Supplement Fig. S2) and found that this was not the case. In TCGA basal-like breast cancer, the ELF3 amplification proportion was not higher than that in other subtypes but was only 3.5%, which was lower than the total level (7.3%). These data demonstrate that the upregulation of ELF3 in BRCA1-associated breast cancer is not caused by copy number amplification.

Furthermore, we investigated whether ELF3 expression could be a biomarker for breast cancer prognosis. Consistent with our hypothesis, higher ELF3 expression is correlated with worse prognosis in TNBC and basal-like breast cancer (Fig. 2G). In the Her2 subtype, higher ELF3 expression is also related to shorter overall survival, consistent with previous reports that as a Her2 target gene, ELF3 may help promote tumor invasion (20). Interestingly, ELF3 expression shows less or no prognostic prediction value in luminal A and luminal B subtypes, suggesting that ELF3 plays a more powerful role in TNBC and basal-like breast cancer. These data imply that in BRCA1-associated breast cancer, ELF3 tends to be upregulated and plays a “driver” role, promoting tumor progression and leading to a worse prognosis.

ELF3 upregulation is induced by replication stress via the ATR-Chk1-E2F axis and by BRCA1 deficiency via GATA3 transcription

We next focused on the mechanisms of ELF3 upregulation in the presence of replication stress and BRCA1 deficiency. Consistent with the HU RNA-seq analysis, our single-cell RNA-seq data showed that in both BRCA1 WT and mutant carriers, the levels of replication stress and ELF3 expression showed a significant positive correlation (Fig. 3A). ELF3 was also upregulated by HU-induced replication stress in MCF10A cells (Fig 3B). Since ATR is the key kinase that responds to replication stress, we treated cells with ATRi and found that it could efficiently block ELF3 upregulation induced by HU, while ATMi had no such effect (Fig. 3B). Moreover, downstream of ATR, ELF3 expression upregulation depended on Chk1 activation but not Chk2 activation (Fig. 3C). It has been reported that the transcriptional activation induced by replication stress-ATR-Chk1 pathway largely depends on E2F transcription factor activation (22). Under replication stress, ATR-Chk1-mediated phosphorylation of E2F6 blocks its inhibitory function on E2F and activates E2F transcription. Thus, we overexpressed E2F6 and found that E2F6 blocks the upregulation of ELF3 under HU treatment conditions (Fig. 3D), validating that ELF3 upregulation relies on E2F6 inactivation and E2F de-suppression. Moreover, utilizing ENCODE ChIP-seq data, we found that the ELF3 promoter is indeed bound by E2F1 (Fig. 3E). These data demonstrate that ELF3 is upregulated by the ATR-Chk1-E2F axis under replication stress conditions.

ELF3 upregulation is induced by replication stress via the ATR-Chk1-E2F axis and by BRCA1 deficiency via GATA3 transcription.

(A) Spearman correlation analysis of ELF3 expression levels and replication stress scores in normal mammary tissue from noncarriers (left) and BRCA1 mutation carriers (right).

(B) Western blot and RT–qPCR results of MCF10A cells treated with the indicated drugs for 12 h. Drug concentration: HU 1 mM, ATRi (VE821) 10 μM, ATMi (KU5593) 10 μM.

(C) Western blot and RT–qPCR results of MCF10A cells treated with the indicated drugs for 12 h. Drug concentrations: HU 1 mM, ATRi (VE821) 10 μM, Chk1i (GDC-0575) 50 μM, Chk2i (BML-277) 10 μM.

(D) Western blot results of MCF10A cells transfected with the empty vector or Myc-E2F6 and treated with DMSO or 1 mM HU for 12 h.

(E) ChIP-Seq enrichment signal (green plot, ENCODE ENCFF858GLM; data are representative of two independent experiments) and the location of detected peaks (black line, ENCODE ENCFF692OYJ; irreproducible discovery rate (IDR) thresholded peaks) of E2F1 in MCF7 cells from the ENCODE database.

(F) ELF3 expression changes as BRCA1 deficiency duration increases. ELF3 expression levels are indicated by the CPM of the RNA-Seq results (means ± SD, n=3).

(G) Western blot and RT–qPCR results of MCF10A cells transfected with siNC or siBRCA1 for 48 h.

(H) RT–qPCR results of MCF10A cells transfected with siNC or siGATA3 for 48 h.

(I) ChIP-Seq enrichment signal (yellow plot up, ENCODE ENCFF384CPN; yellow plot down, ENCODE ENCFF342GNN. Data are representative of two independent experiments) and the location of detected peaks (black line up, ENCODE ENCFF352QVM; black line down, ENCODE ENCFF437NQS; both are irreproducible discovery rate (IDR) thresholded peaks) of GATA3 in MCF7 cells from two independent studies (ENCSR423RTK and ENCSR000BST) from the ENCODE database. The motif sequences were obtained from the JASPAR database.

(J) RT–qPCR results of MCF10A cells transfected with the indicated siRNA for 48 h.

(K) Western blot and RT–qPCR results of MCF10A cells transfected with siNC or siBRCA1 for 24 h followed by treatment with DMSO or 1 mM HU for 12 h.

(L) RT–qPCR results of MCF10A cells transfected with siNC or siGATA3 for 24 h followed by treatment with 1 mM HU for 12 h. In all panels of RT–qPCR results, data are presented as the mean ± SD, n=3, *, p < 0.05; **, p < 0.01; ***, p < 0.001; ns, no significance, by paired two-tailed Student’s t test.

Then we investigate the mechanism of ELF3 upregulation induced by BRCA1 deficiency. ELF3 expression gradually increased during 10 days of DOX treatment (Fig. 3F). ELF3 upregulation was further confirmed by siRNA-mediated BRCA1 knockdown in MCF10A cells to strictly exclude the unknown influence of the Tet-on system (Fig. 3G). BRCA1 has transcription regulation functions by interacting with other transcription factors to bind to gene promoters and regulate gene expression (23). Therefore, we explored transcription factors associated with BRCA1, including GATA3 (24) and Myc (25), for whether the binding motifs of these transcription factors appear in the ELF3 promoter sequence utilizing the JASPAR database (Supplement Fig. S3A). We found that E2F1 was present in the prediction list with a high relative score, consistent with our results. The prediction results show that the motif of GATA3 appeared in the ELF3 promoter multiple times with an overall high relative score. Indeed, knockdown of GATA3 in MCF10A cells also strongly induced ELF3 expression (Fig. 3H). Similarly, we found multiple GATA3 binding peaks at the promoter of ELF3 in two independent ChIP-seq datasets from the ENCODE database, consistent with two of the GATA3 binding sequences predicted in the JASPAR database (Fig. 3I). These data suggest that ELF3 is a GATA3 downstream target gene. Importantly, when both BRCA1 and GATA3 were knocked down, there was no significant superimposed effect on ELF3 upregulation (Fig. 3J and Supplement Fig. S3B), implying that BRCA1 and GATA3 act in the same pathway for ELF3 induction.

Notably, we found that and BRCA1 knockdown and HU treatment have a superimposed effect on upregulating ELF3 expression levels (Fig. 3K and Supplement Fig. S3C). Furthermore, GATA3 knockdown and HU treatment also had a significant superimposed effect in upregulating ELF3 expression (Fig. 3L and Supplement Fig. S3D). These data validate that the BRCA1-GATA3 axis and replication stress regulates ELF3 expression independently.

ELF3 helps suppress excessive genomic instability

Next, we investigated the impact of ELF3 upregulation under replication stress conditions and BRCA1 deficiency. Based on the association between ELF3 and the worse prognosis in BRCA1-associated breast cancer patients (Fig. 2G), we hypothesized that upregulation of ELF3 could help cells deal with replication stress and suppress excessive genomic instability, facilitating cancer initiation and cancer evolution. As expected, ELF3 knockdown greatly increased the sensitivity of MCF10A cells to HU and cisplatin (Fig. 4A and B, Supplement Fig. S4A). This finding suggests that ELF3 can suppress genomic instability and help normal cells deal with replication stress. Furthermore, ELF3 knockdown in HCC1937 and SUM149PT cells, both of which are BRCA1 mutant human breast cancer cell lines, also caused a significant decline in cell proliferation (Fig. 4C and D, Supplement Fig. S4B and C), suggesting that ELF3 is essential for BRCA1 mutant breast cancer cell survival. Moreover, HCC1937 and SUM149PT cells became more sensitive to HU and cisplatin when ELF3 was deficient (Fig. 4E and F, Supplemental Fig. S4D and E). Since PARP inhibitors can induce replication fork stalling by PARP1 trapping and lead to replication stress in BRCA1-deficient cells, we also tested olaparib sensitivity in HCC1937 cells and obtained similar results (Supplement Fig. S4F). These results demonstrate that in BRCA1 mutant breast cancer cells, ELF3 can help suppress excessive genomic instability.

ELF3 helps suppress excessive genomic instability.

(A and B) Cell viability curve of MCF10A cells transfected with siNC or siELF3 for 48 h, treated with the indicated concentration of drugs (HU: μM; cisplatin: nM) and measured by the CCK-8 assay (mean ± SD, n=3, two-way ANOVA, **, p < 0.01; ***, p < 0.001; ****, p < 0.0001).

(C and D) Cell viability of HCC1937 (C) and SUM149PT (D) cells transfected with siNC or siELF3 for 5 days detected by the CCK-8 assay (mean ± SD, n=4, paired two-tailed Student’s t test, **, p < 0.01; ***, p < 0.001; ****, p < 0.0001).

(E and F) Cell viability curve of HCC1937 cells transfected with siNC or siELF3 for 48 h, treated with the indicated concentration of drugs (HU: μM; cisplatin: nM) and measured by the CCK-8 assay (mean ± SD, n=3, two-way ANOVA, **, p < 0.01; ***, p < 0.001; ****, p < 0.0001).

(G and H) Correlation of ELF3 expression levels and cell sensitivity of the indicated drugs in the CellMiner (G) and GDSC (H) databases.

(I) Nude mouse tumors of SUM149PT cells infected with shCtrl of shELF3 lentivirus (mean ± SD, unpaired two-tailed Student’s t test, *, p < 0.05).

In addition, utilizing the CellMiner database, we found that among nearly sixty diverse human cancer cell lines, lower ELF3 expression is correlated with higher drug sensitivity to cisplatin and etoposide (Fig. 4G). Consistent results were obtained using the GDSC database (Fig. 4H, Supplement Fig. 4G). Moreover, in nude mice, SUM149PT tumors with ELF3 deficiency were significantly smaller than those in the control group (Fig. 4I). These data indicate that ELF3 can help suppress excessive genome instability in the presence of replication stress, helping normal cells tolerate replication stress and subsequent DNA damage, allowing the cells to transform into malignant tumors. During cancer progression, BRCA1 mutant cancer cells gradually become reliant on high ELF3 expression to preserve a certain degree of genomic stability, making ELF3 a promising therapeutic target and biomarker in BRCA1-associated breast cancers.

ELF3 helps maintain the stability of DNA replication

We then investigated the mechanism by which ELF3 preserves genomic stability. ELF3 deficiency leads to more endogenous DNA damage in MCF10A cells (Supplement Fig. S5). In HCC1937 cells, knockdown of ELF3 resulted in a significant increase in spontaneous γH2AX foci (Fig. 5A), indicating that ELF3 deficiency leads to more endogenous DNA damage in BRCA1-deficient breast cancer cells. Moreover, with HU treatment and release for repair, ELF3 deficiency resulted in a significant increase in the number of γH2AX and 53BP1 foci per cell and the proportion of cells with damage foci (Fig. 5B-D). Under replication stress conditions, γH2AX and 53BP1 foci represent DNA double-strand breaks formed by collapsed replication forks (26). These data suggest that ELF3 deficiency results in a defective ability to repair stalled replication forks; thus, more DNA damage is generated. The resulting upregulation of ELF3 expression during replication stress helps cells better manage stalled replication forks and prevents excessive DNA damage.

ELF3 helps maintain the stability of DNA replication.

(A) Spontaneous γH2AX foci of HCC1937 cells transfected with siNC or siELF3 for 72 h (mean ± SEM, n > 100, unpaired two-tailed Student’s t test, **, p < 0.01; ****, p < 0.0001).

(B and C) γH2AX and 53BP1 foci of MCF10A cells transfected with siNC or siELF3 for 48 h and treated with 2 mM HU for 4 h followed by release for 20 h (mean ± SEM, n > 100, unpaired two-tailed Student’s t test, *, p < 0.05; **, p < 0.01; ***, p < 0.001).

(D) Proportions of cells with γH2AX (⩾ 3) and 53BP1 foci from (B and C) (mean ± SD, unpaired two-tailed Student’s t test, *, p < 0.05; **, p < 0.01; ***, p < 0.001).

(E) Enrichment plot of DNA replication pathways in MCF10A ELF3 knockdown RNA-seq.

(F) RT–qPCR results of MCF10A cells transfected with siNC or siELF3 for 72 h (mean ± SD, n=3, paired two-tailed Student’s t test, *, p < 0.05; **, p < 0.01; ns, no significance).

(G) Replication fork velocity indicated by DNA fiber assay of MCF10A cells transfected with the indicated siRNA for 72 h (mean ± SD, n > 80, unpaired two-tailed Student’s t test, ****, p < 0.0001).

(H) Replication fork symmetry indicated by the DNA fiber assay of MCF10A cells transfected with the indicated siRNA for 72 h (mean ± SD, Mann–Whitney test, ***, p < 0.001; ****, p < 0.0001).

(I) Replication fork stability of MCF10A cells transfected with the indicated siRNA for 72 h (mean ± SD, n > 100, unpaired two-tailed Student’s t test, **, p < 0.01; ****, p < 0.0001).

To further explore how ELF3 contributes to replication stress management, RNA-seq on ELF3-knockdown MCF10A cells was performed. Gene set enrichment analysis (GSEA) showed that multiple pathways associated with DNA replication were significantly downregulated in ELF3-deficient cells (Fig. 5E), which was further verified by RT–qPCR (Fig. 5F). These data indicate that DNA replication pathways are affected by ELF3 deficiency.

DNA fiber assay was performed to more closely investigate the effects of ELF3 deficiency on DNA replication. ELF3 deficiency significantly slowed down the DNA replication progression (Fig. 5G), with significantly reduced symmetry of replication forks (Fig. 5H), suggesting that more replication forks were stalled or unstable. Moreover, with HU treatment, ELF3 knockdown undermined the stability of stalled replication forks (Fig. 5I). Notably, double-knockdown of ELF3 and BRCA1 led to a further decrease in replication fork stability (Fig. 5I). These data reveal that ELF3 expression upregulation when BRCA1 is deficient can help cells deal with replication stress, while simultaneous ELF3 deficiency leads to intolerable levels of genomic stability.

ELF3 is a core transcription factor inducing LP genes expression under BRCA1 deficiency

Finally, we explored the possible reasons for the tissue specificity of BRCA1-associated tumors. It has been reported that BRCA1 deficiency disturbs the development and differentiation of mammary tissue and causes abnormal expansion of the LP population (3, 8, 27). We hypothesized that BRCA1 deficiency might affect the normal differentiation process of LPs by disturbing the transcriptional profile, thus leading to abnormal expansion and transformation of LPs. We analyzed the time-course RNA-seq data using a published signature gene set from LPs (3) and found that over time with BRCA1 deficiency, normal mammary cells gradually show a higher signature score of LP gene expression (Fig. 6A). Similarly, using the sets of genes up-and downregulated in the LP population compared to in other mammary cell populations obtained from the data of Pellacani et al. (28), we found that genes upregulated in LPs also tended to be upregulated in BRCA1-deficient conditions, whereas genes downregulated in LPs tended to be downregulated (Fig. 6B). These results suggest that BRCA1 may broadly affect the transcription profile of mammary cells, and therefore, when BRCA1 is deficient, LPs may tend to maintain a dedifferentiated state, leading to the abnormal expansion and tumorigenic properties of these cells in BRCA1 mutant carriers.

BRCA1 deficiency disturbs the differentiation of LPs.

(A) LP signature score of samples of MCF10A-shBRCA1-Tet-on RNA-seq. The signature score was generated using data from Lim et al. (3).

(B) Barcode plot of LP gene set enrichment in Day 10 vs. Ctrl. LP gene sets were obtained from data from Pellacani et al. (33).

(C) GSEA plots of the transcriptional profile of MCF10A cells with BRCA1 deficiency, MCF10A cells with ELF3 overexpression and normal human LPs. The LP gene set was obtained from the data of Pellacani et al. (33).

(D) Violin plots displaying the distribution of ELF3 expression levels in distinct cell types in normal mammary tissue from noncarriers (left) and BRCA1 mutation carrier (right).

(E) ETS subfamily transcription factor motif enrichment results of LP gene promoters.

(F) Expression levels of ETS subfamily transcription factors in LPs from BRCA1 mutation carrier and noncarrier of the single-cell sequencing data.

(G) ETS subfamily transcription factor enrichment of ATAC-Seq of LPs of BRCA1-deficient mice and WT mice from Bach et al. (37).

Multiple ETS family transcription factors were shown to be involved in human mammary development and differentiation in previous reports. However, the function of ELF3 in LPs has not yet been elucidated. We used GSEA to explore the transcriptional profiles of BRCA1-deficient mammary cells, ELF3-overexpressing mammary cells and human LPs from previous reports (28). Surprisingly, these three transcriptional profiles have significant correlations (Fig. 6C), indicating that ELF3 is not only a core transcription node under BRCA1 deficiency conditions but also, more importantly, that ELF3 plays significant roles in the transcriptional regulation of human LPs. In addition, in our single-cell sequencing data, the expression of ELF3 in LPs is the highest among all mammary cell populations (Fig. 6D and Supplement Fig. S6A). Same results were obtained analyzing the published data from Pellacani et al. (28) (Supplement Fig. S6B). To investigate ELF3 functions in regulating LP genes transcription, we used SEA (simple enrichment analysis) of MEME to perform motif enrichment of ESE and ELF subfamilies transcription factors in promoters of LP genes, as these transcription factors are more closely related to ELF3 in the ETS family, and several, such as ELF5 and EHF, have been reported to regulate LPs (28-30). ELF3 was at the top of the list (Fig. 6E), demonstrating that ELF3 has higher affinity for the promoters of LP genes. Moreover, our single-cell data revealed that ELF3 expression levels were the highest in LPs compared to other members of the ESE and ELF subfamilies (Fig. 6F), including ELF5, which has been reported to play vital roles in LP differentiation. In addition, we analyzed the ATAC-seq data in Brca1-deficient mouse LPs from Bach et al. (31). The ATAC-seq compared the chromatin accessibility of LPs in Brca1/p53-deficient mice with WT mice, and motif enrichment analysis of transcription factors was performed in regions with increased accessibility of Brca1-deficient mouse LPs. ELF3 was among the top genes (Supplement Fig. S6C) and is significantly enriched among all ETS transcription factors (Fig. 6G). These data further confirm our hypothesis that ELF3 plays an important role in the transcriptional regulation of LPs.

Discussion

In this study, we revealed that LPs have the trait of higher replication stress, an important factor for their potential to transformation. The difference of LPs from other cell populations in normal mammary tissue has received considerable attention in the field. Research in mouse mammary tissue has revealed that the overall luminal epithelial cell population is enriched in mitotic, cell cycle and DNA repair-related pathways under estrogen-and progestogen-stimulated conditions relative to the basal and stromal cell populations (32). Subsequently, Casey et al. (33) found that LPs of mouse mammary tissue are enriched in DNA repair and cell division pathways. A recent preprint from the same group (34) continued the analysis of mouse tissue sequencing data and found that in LPs, DNA repair and replication pathways are significantly upregulated, indicating that LPs are more capable of DNA damage repair and thus more tolerant to BRCA1 deficiency. Furthermore, Nolan E et al. (5) found that RANK+ LPs are highly proliferative and prone to DNA damage. These studies did not reveal the reason for LP tendency to transform. In our study, we focus on the characteristic of LPs compared to other cell populations and using sequencing data from normal human mammary tissues (3), we reveal that the DNA replication pathways and the cellular response to replication stress pathways are significantly upregulated in LPs compared to those in other cell populations. We argue that high levels of replication stress may be the major source of DNA damage in LPs, resulting in DDR gene upregulation in these cells, and the high proliferation feature of LP could be the main reason for the high replication stress.

Replication stress is a hallmark of cancer, fueling cancer evolution by increasing genomic instability (35). LPs have inherently higher replication stress levels, and when BRCA1 is deficient, replication stress is further elevated during the transformation from normal LPs to tumor cells, suggesting that replication stress is an important driver of BRCA1-deficient tumorigenesis and the suppression of replication stress could be a new approach for preventive intervention in BRCA1 mutation carriers with the advantages of being less damaging than prophylactic resection. Moreover, we found that both replication stress and BRCA1 deficiency itself can induce high expression of ELF3, which also plays an important role in the tendency of LPs to transform with BRCA1 deficiency. First, ELF3 can help suppress excessive genomic instability, which is crucial for cell tolerance and survival with BRCA1 deficiency. We speculate that high ELF3 expression can help LPs maintain moderate levels of genomic instability that allow cells to proliferate and is sufficient to fuel cancer evolution, including subsequent events such as p53 mutation and BRCA1 loss of heterozygosity (LOH). Second, ELF3 upregulation can boost LP gene transcription, leading to LP dedifferentiation and expansion. Finally, as an EMT promoter in mammary tissue, ELF3 can accelerate the transformation of LPs to malignant cells. Therefore, high levels of replication stress and ELF3 expression are important factors resulting in the high tendency for LPs to transform (Fig. 7). A key feature of BRCA1-associated tumors is the tissue specificity, the mechanism of which has long been a research focus. The transcriptional regulation functions of BRCA1 are critical for the development and differentiation of normal mammary tissue. LPs with abnormal expansion have been observed in BRCA1 mutation carriers. Several important differentiation pathways, such as WNT, Notch and Hedgehog, were downregulated in these abnormal LPs (8), suggesting that BRCA1 deficiency may lead to a differentiation blockade in LPs and, therefore, abnormal expansion. Our RNA-seq data revealed that the transcriptional profile of BRCA1-deficient MCF10A cells has a striking similarity to that of LPs in normal human mammary tissue, and ELF3 might be a core mediating transcription factor. This finding suggests that BRCA1 has a broad regulatory effect on gene expression in normal mammary cells and has specific functions in LPs; thus, its deficiency causes LPs to maintain a dedifferentiated state, leading to abnormal expansion of this cell population, and ELF3 may be a core transcription factor inducing these changes. In addition, our study reveals another layer of the mechanism underlying the tissue specificity of BRCA1-associated tumors—the capability of BRCA1 to regulate ELF3 expression. ELF3 is an epithelial-specific transcription factor, but the functions of ELF3 in different epithelial tissues vary. For example, ELF3 promotes tumor development in breast, prostate, and lung cancers (10, 36, 37), whereas it suppresses tumorigenesis in bladder and oral squamous cancers (38, 39). In addition, ELF3 itself plays an important role in the development and differentiation of normal mammary tissue; thus, dysregulation of ELF3 may disturb these processes. Therefore, we suggest that replication stress caused by BRCA1 deficiency leads to upregulation of ELF3 expression in LPs, thus promoting the tissue-specific oncogenic functions of ELF3 and breast tumorigenesis.

Replication stress inducing ELF3 upregulation promotes BRCA1-deficient breast tumorigenesis.

LPs have higher replication stress levels under normal conditions (left) and thus are more likely to exceed the tumorigenesis threshold under BRCA1 deficiency (right). Replication stress and BRCA1 deficiency results in further ELF3 upregulation (top right). ELF3 can suppress excessive replication stress and help LPs maintain moderate levels of genomic instability, which is tolerable for cells to proliferate and enough to fuel cancer evolution (a). In addition, ELF3 upregulation can boost LP gene transcription, leading to LP dedifferentiation and expansion (b). Finally, as an EMT promoter in the mammary tissue, ELF3 can accelerate the transformation process from LPs to malignant cells (c).

This study also has some limitations. First, we used the MCF10A cell line, which is widely used for investigations into normal human mammary epithelial cells with the advantage of being easy to manipulate and modify, and compared to patient-derived cells, unknown genetic backgrounds in addition to BRCA1 mutations are avoided. However, the transcriptional profile and epigenetic characteristics of this cell line are distinct from those of normal human mammary cell populations and LPs (28). Although our data show that MCF10A cells can reflect the actual functions of BRCA1 in terms of maintaining genomic stability (Fig. 1B and C) and our RNA-seq data demonstrate enrichment in the known transcriptional target genes of BRCA1, the effect of BRCA1 defects in LPs needs to be further verified in primary human mammary cell populations. Second, the higher replication stress levels in LPs may lead to altered expression or modifications of molecules other than ELF3, affecting the development and progression of BRCA1-associated breast cancer, which also needs to be further explored. In addition, there are likely other DEGs in our BRCA1-deficient RNA-seq that play important roles in the progression of BRCA1-associated breast cancer, and these genes also have potential research value.

Taken together, we revealed that LPs have a trait of higher replication stress, endowing them with the potential for transformation with BRCA1 deficiency. Moreover, the levels of replication stress gradually increase during tumorigenesis, indicating that replication stress is a driving factor of BRCA1-associated malignant transformation. Mechanistically, ELF3 is significantly upregulated upon replication stress and BRCA1 deficiency, which can help suppress excessive genomic instability by stabilizing the DNA replication process, empowering LPs to tolerate and to transform with BRCA1 deficiency. Notably, BRCA1 deficiency causes a tendency for cells to display characteristics of the LP transcriptional profile, and ELF3 is a core transcription factor mediating this process. Our study reveals why BRCA1 deficiency is prone to result in tumorigenesis in LPs and elucidates the key role of replication stress and ELF3 during this process, and suggests promising targets for BRCA1-associated breast cancers.

Materials and Methods

Cell culture

MCF10A cells were cultured in DMEM/F12 supplemented with 5% horse serum, 1% penicillin/streptomycin, 20 ng/mL EGF, 10 μg/mL insulin, 0.5 μg /mL hydrocortisone and 100 ng/mL cholera toxin. For MCF10A-shBRCA1-Tet-on cells, except for the replacement of horse serum with 8% Tet-on system approved certified fetal bovine serum, all other components remain the same. HEK293T and SUM149PT cells were cultured in DMEM supplemented with 10% fetal bovine serum and 1% penicillin/streptomycin. HCC1937 cells were cultured in RPMI1640 supplemented with 10% fetal bovine serum and 1% penicillin/streptomycin.

Reagents and plasmids

Reagents (antibodies and chemicals) used in this study were listed in Table S1. For the construction of Tet-on-shBRCA1-hygro plasmid, shBRCA1 sequence was subcloned into the Tet-on-pLKO-hygro vector (SinaSun). For ELF3 knockdown by lentivirus in SUM149PT, shELF3 and shCtrl sequence was subcloned into the pLKO.1-puro vector. The shRNA sequences used in this study are listed in Table S2. For ELF3 overexpression by lentivirus in MCF10A, human ELF3 was subcloned into the pCDH-CMV-MCS-EF1-puro vector, which was a kind gift from Yin lab. Myc-tagged human E2F6 was subcloned into the pDEST-N-Myc vector. PCR primers used in this study were listed in Table S3.

Transfection of siRNA and infection of shRNA

All the siRNAs were from GenePharma. For siRNA transfection, cells were seeded in 6-well plate or 35 mm2 dish at the day before transfection. 100 μL OPTI-MEM, 40 pmol siRNA and 3 μL Lipofectamine RNAiMAX reagent were mixed and added to the cells. After 48-72 h, cells were collected for subsequent experiments. The siRNA sequences used are listed in Table S3. For shRNA infection, lentivirus was generated by co-transfection of the lentiviral vector (Tet-pLKO.1-shBRCA1, pLKO.1-shCtrl and pLKO.1-shELF3) with envelope plasmids pMD2.G and psPAX2 in HEK293T cells. Lentivirus was collected after 48 h. Cells were infected with lentivirus with polybrene. After 24 h, cells were re-infected and harvested at 72 h for subsequent experiments.

Quantitative PCR (RT-qPCR)

Total RNA was extracted using Trizol reagent (Invitrogen) according to the manufacturer’s guidelines. cDNA was synthesized by reverse transcription PCR using Hifair III 1st Strand cDNA Synthesis SuperMix kit (Yeasen). qPCR was performed using SYBR Green Mix (Yeasen) on ABI StepOnePlus qPCR instrument (ThermoFisher). Each group was performed in triplicate. The relative expression mRNA levels were calculated by relative quantitative method (ΔΔCt). The primers used are listed in Table S5.

Immunofluorescence

Cells were seeded on coverslips and treated as indicated. Cells were than fixed in cold 4% paraformaldehyde for 15 min and permeabilized in 0.25% Triton X-100 solution for 5 min at room temperature. After blocking with 2% BSA for 10 min, the primary antibodies were added onto the coverslips, and then incubated for 1 h at room temperature, followed by secondary antibody incubation for 30 min. Finally, the coverslips were stained with DAPI for 5 min, and images were acquired using the ZEN software.

Immunohistochemistry

All tissue microarray chips were purchased from Shanghai Outdo Biotech Co.Ltd. IHC was performed as described previously, except that the following antibodies were used. The anti-ELF3 rabbit antibody were purchased from SIGMA. After incubation with primary antibody, the HRP-conjugated rabbit secondary antibody (ZSGB-Bio) was applied and IHC slides were observed with Olympus BX51 microscope and Olympus DP73 CCD photographic system.

Colony formation assay

MCF10A-shBRCA1-Tet-on cells were seeded in 6 mm2 plates. 1 μg/mL DOX was added for the DOX group during the whole process of cell growth to knockdown BRCA1 expression. After 14 days, the cells were washed with phosphate-buffered saline (PBS) and stained with 0.1% Coomassie brilliant blue in 10% ethanol for 30 minutes at room temperature. The stained dishes were washed with water, and the colonies were counted.

DNA fiber

The DNA fiber assay was performed as described previously with slight modifications (40). Briefly, cells transfected with siRNA were labeled by 25 μM IdU (Yuanye Bio-Technology) for 20 min and subsequently 200 μM CldU (Abcam) for 20 min. Then cells were treated with 4 mM HU for 4 h for replication fork stability assay or collected for replication fork speed and symmetry assay. Cells were resuspended in lysis buffer (200 mM Tris-HCl, pH 7.5, 50 mM EDTA and 0.5% SDS) to extract DNA. Cell lysates were then dripped onto a glass slide and DNA fibers were stretched along the slide after 3 minutes incubation. The slides were air-dried and fixed in 3:1 methanol/acetic acid solution for 10 min, followed by 2.5 M HCl treatment for 1 h, then blocked in 3% BSA PBST buffer for 20 min. DNA fibers were incubated with anti-BrdU antibody (Abcam) binding CldU and anti-BrdU antibody (BD) binding IdU at 37°C for 1.5 h. After washing with PBST, the slides were incubated with Alexa Fluor 488- and Alexa Fluor 594-conjugated secondary antibody (Invitrogen) for 45 minutes. Finally, the slides were mounted in mounting reagent (Solarbio) and DNA fibers images were acquired using ZEN software.

Comet assay

The comet assay was performed as described previously with slight modifications (41). Briefly, cells were collected by trypsin digestion and resuspended in PBS. 0.5% normal melting agarose (NMA) was added to glass slides and quickly covered with coverslips. The slides were placed at 4 °C for 10 min. The coverslips were removed carefully. The cell suspensions were mixed gently with equal volume of 1% low melting agarose (LMA) and quickly added onto the solidified NMA, quickly covered with coverslips. The slides were placed at 4 °C for 10 min. After removing the coverslips, the slides were incubated overnight in the fresh cell lysis buffer at 4 °C overnight. Then the slides were rinsed with ddH2O and electrophoresis at 4°C for 30 minutes at 20 V in neutral electrophoresis buffer. Then the slides were stained with 5 μg/ml of propidium iodide for 10 min. The comet was observed using a fluorescence microscope (Zeiss AxioCam 503 color). The pictures were analyzed using CASP software (http://casplab.com/).

Cell survival (CCK-8) assay

Cell survival assay was performed using CCK-8 kits (meilunbio, MA0218-3) according to the manufacturer’s protocols. Cells were seeded in 96-well plates. After 12 h, medium containing indicated concentration of drugs was changed and cells were incubated for 3-5 days. DMEM containing 0.1% DMSO was used as the control group. CCK-8 solution was diluted in 10% concentration and added to wells, and the cells were incubated for 45 min-1.5 h. DMEM containing 10% CCK-8 was used as the blank group. The absorbance at 450 nm was detected by a microplate reader. Cell survival was calculated according to the formula: (ODdrug – ODblank) /(ODcontrol – ODblank).

RNA sequencing

Total RNA of cells was extracted with Trizol reagent. RNA sequencing was performed by Novogene (China). The resulting data was mapped to human reference genome (hg38) and gene counts were calculated for each sample using STAR (v2.6.1) (42) with the parameter “--quantMode GeneCounts”. To identify differentially expressed genes (DEGs), genes with low expression levels (average counts per million reads < 0.05) were filtered out. DESeq2 was employed to detect DEGs with |log2 fold change| > 2 and FDR < 0.001. The raw RNA sequencing data have been deposited into GEO under the accession number GSExxxxxx. PCA analysis was performed using the “prcomp” function of stats (v3.6.1) R package.

Analysis of TCGA breast cancer and METABRIC data

TCGA (https://www.cancer.gov/tcga) breast cancer and METABRIC (18) data were downloaded using the cgdsr R package. For TCGA data, RSEM (batch normalized from Illumina HiSeq_RNASeqV2) was used as the expression levels of the genes. For METABRIC data, mRNA expression log intensity levels (Illumina Human v3 microarray) were used as the expression levels of the genes. BRCA1 associated cancers were identified as cancers with BRCA1 mutations, heterozygous loss or homozygous deletion. Pearson correlation coefficient between ELF3 expression and BRCA1 expression was calculated. Subtypes and IntClust types of breast cancers were identified according to the clinical data of TCGA and METABRIC databases.

Chip-Seq analysis of ENCODE databases

The Chip-Seq data were downloaded from the ENCODE database (https://www.encodeproject.org/) (43, 44) with the following identifiers: ENCFF858GLM, ENCFF692OYJ, ENCFF384CPN, ENCFF342GNN, ENCFF352QVM, and ENCFF437NQS. Images were generated using IGV (v2.7.2).

Drug sensitivity analysis of Cell Miner and GDSC data

Cell line drug sensitivity data and mRNA expression data were downloaded from the CellMiner (45, 46) and GDSC (47) databases. Pearson correlation coefficient between ELF3 expression and drug sensitivity was calculated for the indicated drugs.

Replication stress score and luminal progenitor score analysis

For replication stress score calculation, human mammary subgroups gene expression matrixes were obtained from the datasets of Lim et al. (3) (GSE16997). Replication stress associated pathways were selected referring to Dreyer et al. (13). and gene lists were downloaded from the GSEA website (www.gsea-msigdb.org). For each pathway, replication stress scores were calculated according to the formula below, where eg is the expression of the gene in different mammary subgroups, and ng is the number of genes in this pathway.

For luminal progenitor score calculation, the LP signature gene set with average log-fold change was obtained from data of Lim et al. (3), and log CPM matrix of MCF10A-shBRCA1-Tet-on DOX 10 days vs Ctrl RNA-seq data was used. For each sample, LP scores were calculated according to the formula below, which refers to Lim et al. (3), where xg is the log-fold-change for the gene from LP gene set, and yg is the log cpm for the same gene in BRCA1 knockdown RNA-seq.

GSEA analysis and barcode plot

GSEA analysis was performed using GSEA software (v4.1.0) (48, 49) and clusterProfiler (v3.14.3) (50). For the customized gene set enrichment analysis of the transcriptional profiles of BRCA1 KD, ELF3 OE and LP, BRCA1 KD data were from MCF10A-shBRCA1-Tet-on DOX 10 days vs Ctrl RNA-Seq data and ELF3 OE data were from MCF10A OE ELF3 vs Ctrl RNA-seq data. LP gene set was obtained from 2016 cell reports, and OE ELF3 gene sets were generated using MCF10A OE ELF3 vs Ctrl RNA-seq data, where the log-fold-change cutoff was 1.3 and P value cutoff was 0.01. For the barcode plot, LP gene sets were obtained from data of Pellacani et al. (28), and log-fold-change of all genes from MCF10A-shBRCA1-Tet-on DOX 10 days vs Ctrl RNA-seq data was used for the enrichment analysis. P value was calculated using the “roast” function from limma (v3.42.2) (51).

Motif enrichment analysis

Motif matrixes of transcription factors were downloaded from the JASPAR database (52). LP gene list was obtained according to the data of Pellacani et al. (28) and LP genes are identified to significantly upregulated in LPs compared with all three other subgroups (BC, LC and SC). Promoter sequences of LP genes were extracted from the genome from 3000 bp upstream to 200 bp downstream of the TSS sites. Motif enrichment analysis was performed using SEA (Simple Enrichment Analysis) of the MEME Suite (53).

Single-cell sequencing data analysis

Single-cell RNA-seq (scRNA-seq) expression profiles of breast cancer tissue and normal mammary tissue from four BRCA1 mutation carriers, as well as normal mammary tissue from three non-carriers were generated using the 10x Genomics Chromium platform and NovaSeq 6000 platform. The detail information of sample processing, scRNA-seq data processing and quality control were described in our previous study (8). Briefly, scRNA-seq data analyses was completed utilizing the 10× Genomics Cell Ranger software suite. Principal component analysis (PCA) (twenty-five principal components) and k-means algorithm (k was set as 9.0) were used to cluster cells. We used a series of cell specific markers to identify the cell type of each cluster, as described in our previous study1. Among these, luminal progenitors cluster was identified based on the expression of KRT8/18+, CD24+, GABRP+, KIT+, and ALDH1A3+. Triple negative tumor cell population was identified based on tumor sample specific cluster, ER/PR/ HER2 expression coincident with the clinical records and PAM50 subtypes analysis. Replication stress score of each cell was calculated based on the sum expression level of genes involving in replication stress-related pathways referring to Dreyer et al. (13). In this study, we selected luminal lineage cells and tumor cells from a BRCA1-mutated triple-negative breast cancer patient to construct tumorigenesis trajectories with the R package Monocle 2 with standard settings.

Animal experiments

5×106 SUM149PT tumor cells were subcutaneously injected into the flanks of female BALB/c nude mice. Tumor growth was measured using a caliper every three days, and the tumor volume was calculated using the formula: tumor volume = length ×width2/2.

Statistics Analysis

Statistical analysis was performed using GraphPad Prism (GraphPad Software). The tests used included the paired and unpaired two-tailed Student’s t test, the Mann-Whitney U test, one-way and two-way ANOVA, and Pearson’s correlation statistical test. A P value of less than 0.05 was considered statistically significant. Details can be found in figure legends.

Study approval

The normal human breast cancer tissue and breast cancer samples used for single-cell RNA-seq in this study were collected from Peking University Cancer Hospital. This study was carried out in accordance with the ethical principles of the Declaration of Helsinki and was approved by the Research and Ethics Committee of Peking University Cancer Hospital. Written informed consent was obtained from all participants prior to participation. Animal experiments in this study were performed in accordance with the Guidelines of Peking University Animal Care and Use Committee.

Supporting information

eLife Assessment

This important study reveals ELF3 as a putative candidate driver of luminal progenitor (LP) transformation. Up-regulation of ELF3 during replicative stress conditions and in BRCA1 deficient cells may permit cell proliferation by suppressing genome instability. While the hypothesis is compelling, the experimental support is still incomplete, as it does not adequately demonstrate the role of ELF3 in LP cells per se. The mechanistic underpinnings by which ELF3 promotes cell tolerance to DNA damage were not fully explored either. With improvements, the work has the potential to enhance our understanding of how BRCA1 deficiency fuels LP transformation and thereby breast tumorigenesis.

Reviewer #1 (Public Review):

The authors set out to define the molecular basis for LP as the origin of BRCA1-deficient breast cancers. They showed that LPs have the highest level of replicative stress, and hypothesise that this may account for their tendency to transform. They went on to identify ELF3 as a candidate driver of LP transformation and showed that ELF3 expression is up-regulated in response to replicative stress as well as BRCA1 deficiency. They went on to show that ELF3 inactivation led to a higher level of DNA damage, which may result from compromised replicative stress responses.

While the manuscript supports the interesting idea wherein ELF3 may fuel LP cell transformation, it remains obscure how ELF3 promotes cell tolerance to DNA damage. Interestingly the authors proposed that ELF3 suppresses excessive genomic instability, but in my opinion, I do not see any evidence that supports this claim. In fact, one might think that genomic instability is key to cell transformation.

Reviewer #2 (Public Review):

Summary:

The manuscript focuses on a persistent question of why germline mutations in BRCA1 which impair homology-directed repair of DNA double-strand breaks predispose to primarily breast and ovarian cancers but not other tissues. The authors propose that replication stress is elevated in the luminal progenitor (LP) cells and apply the gene signature from Dreyer et al as a measure of replication stress in populations of cells selected by FACS previously (published by Lim et al.) and suggest an enrichment of replication stress among the LP cells. This is followed by single-cell RNA seq data from a small number of breast tissues from a small number of BRCA1 mutation carriers but the pathogenic variants are not listed. The authors perform an elegant analysis of the effects of BRCA1 knockdown in MCF10A cells, but these cells are not considered a model of LP cells.

Overall, the manuscript suffers from significant gaps and leaps in logic among the datasets used. The connection to luminal progenitor cells is not adequately established because the models used are not representative of this population of cells. Therefore, the central hypothesis is not sufficiently justified.

Strengths:

The inducible knockdown of BRCA1 provided compelling data pointing to an upregulation of ELF3 in this setting as well as a small number of other genes. It would be useful to discuss the other genes for completeness and explain the logic for focusing on ELF3. Nonetheless, the connection with ELF 3 is reasonable. The authors provide significant data showing a role for ELF3 in breast epithelial cells and its role in cell survival.

Weaknesses:

The initial observations in primary breast cells have small sample sizes. The mutations in BRCA1 seem to be presumed to be all the same, but we know that pathogenic variants differ among individuals and range from missense mutations affecting interactions with one critical partner to large-scale truncations of the protein.

The figure legends are missing critical details that make it difficult for the reader to evaluate the data. The data support the notion that ELF3 may participate in relieving replication stress, but does not appear to be limited to LP cells as proposed in the hypothesis.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public Review):

The authors set out to define the molecular basis for LP as the origin of BRCA1deficient breast cancers. They showed that LPs have the highest level of replicative stress, and hypothesise that this may account for their tendency to transform. They went on to identify ELF3 as a candidate driver of LP transformation and showed that ELF3 expression is up-regulated in response to replicative stress as well as BRCA1 deficiency. They went on to show that ELF3 inactivation led to a higher level of DNA damage, which may result from compromised replicative stress responses.

While the manuscript supports the interesting idea wherein ELF3 may fuel LP cell transformation, it remains obscure how ELF3 promotes cell tolerance to DNA damage. Interestingly the authors proposed that ELF3 suppresses excessive genomic instability, but in my opinion, I do not see any evidence that supports this claim. In fact, one might think that genomic instability is key to cell transformation.

We greatly appreciate your thorough review and insightful comments on our manuscript. We have taken your feedback seriously and have made several key revisions to address your concerns.

To your primary point about how ELF3 helps cells tolerate DNA damage, we have expanded our discussion to clarify the role of ELF3 in the context of BRCA1 deficiency and high replicative stress. We clarified that while ELF3 may not directly suppress excessive genomic instability, it plays a role in maintaining a balance that prevents catastrophic damage in BRCA1-deficient cells. Both BRCA1 deficiency and increased replication stress induce up-regulation of ELF3, which acts as a transcription factor, and it’s up-regulation leads to up-regulation of the expression of a variety of DNA replication-associated proteins that help to maintain homeostasis in the DNA replication process (Figure 5 E and F). Defects in ELF3 also do lead to disruption of the DNA replication process (Figure 5 G-I). While ELF3 cannot completely eliminate genomic instability, ELF3 essentially maintains genomic instability within a dangerous yet non-lethal range: higher than in normal cells, but not so high as to cause cell death.

This precarious balance can facilitate the transformation of LPs into a malignant state, as you pointed out.

In the revised manuscript, we emphasized that in cells with inherently low replicative stress, such as other non-LP mammary cells, the ELF3-associated mechanism might help cells endure the high replicative stress caused by BRCA1 deficiency without leading to cancerous changes. However, in LP cells, which naturally experience higher replicative stress, this ELF3-related mechanism may make them more susceptible to transformation into cancer cells. This supports our hypothesis that the combination of high replicative stress and BRCA1 deficiency specifically predisposes LP cells to tumorigenesis.

We have modified the working model to make it clearer.

Reviewer #2 (Public Review):

Summary:

The manuscript focuses on a persistent question of why germline mutations in BRCA1 which impair homology-directed repair of DNA double-strand breaks predispose to primarily breast and ovarian cancers but not other tissues. The authors propose that replication stress is elevated in the luminal progenitor (LP) cells and apply the gene signature from Dreyer et al as a measure of replication stress in populations of cells selected by FACS previously (published by Lim et al.) and suggest an enrichment of replication stress among the LP cells. This is followed by single-cell RNA seq data from a small number of breast tissues from a small number of BRCA1 mutation carriers but the pathogenic variants are not listed. The authors perform an elegant analysis of the effects of BRCA1 knockdown in MCF10A cells, but these cells are not considered a model of LP cells.

Overall, the manuscript suffers from significant gaps and leaps in logic among the datasets used. The connection to luminal progenitor cells is not adequately established because the models used are not representative of this population of cells. Therefore, the central hypothesis is not sufficiently justified.

Strengths:

The inducible knockdown of BRCA1 provided compelling data pointing to an upregulation of ELF3 in this setting as well as a small number of other genes. It would be useful to discuss the other genes for completeness and explain the logic for focusing on ELF3. Nonetheless, the connection with ELF 3 is reasonable. The authors provide significant data showing a role for ELF3 in breast epithelial cells and its role in cell survival.

Weaknesses:

The initial observations in primary breast cells have small sample sizes. The mutations in BRCA1 seem to be presumed to be all the same, but we know that pathogenic variants differ among individuals and range from missense mutations affecting interactions with one critical partner to large-scale truncations of the protein.

The figure legends are missing critical details that make it difficult for the reader to evaluate the data. The data support the notion that ELF3 may participate in relieving replication stress, but does not appear to be limited to LP cells as proposed in the hypothesis.

We would like to sincerely thank you for your thorough review and constructive feedback on our manuscript. Your insightful comments and suggestions have been invaluable in guiding our revisions.

(1) Acknowledgment of Data Set Limitations and Additional Analyses: We fully acknowledge the importance of the concerns raised regarding the datasets used in our study. We have supplemented our manuscript with the missing information you pointed out and conducted additional analyses as suggested. These efforts have

(2) Challenges in LP Cell Experiments:

One of the most critical issues you raised was the lack of validation in LP cells, particularly concerning the role of ELF3 in these cells. We are acutely aware of the significance of this point. Following your review, we made extensive efforts to isolate and culture LP cells from both BRCA1-proficient and BRCA1-deficient patient samples. We tried various methods and invested substantial resources, including time, manpower, and materials, to establish a reliable protocol for isolating and cultivating LP cells in vitro. Unfortunately, despite our best efforts, we were unable to obtain a sufficient number of high-quality cells to generate solid and reproducible results.

The challenges we faced included the limited availability of patient tissues and the technical difficulties in consistently obtaining viable LP cells. Given the already extended timeline for the revision of this manuscript, we regretfully decided to forgo further attempts to perform these critical experiments with LP cells. In the revised manuscript, we have explicitly addressed the limitations of our cell models and provided a detailed discussion of the challenges faced in isolating LP cells. Despite these limitations, we believe that the consistency between our results and LP cell sequencing data provides valuable insights and a solid foundation for future studies.

(3) Data Presentation Improvements:

In response to your feedback, we have also made significant improvements to the data presentation in our manuscript. We updated and optimized figure legends and narrative sections to ensure that the data are clearly and accurately conveyed. These changes aim to enhance the readability and comprehensibility of our findings.

We greatly appreciate your valuable feedback, which has significantly contributed to the improvement of our manuscript. Your suggestions have helped us refine our arguments and present a more robust and nuanced interpretation of our data.

Thank you once again for your critical and constructive review. We look forward to your feedback on our revised manuscript.

Recommendations for the authors:

Reviewer #1 (Recommendations For The Authors):

As such, in addition to consolidating the role of ELF3 in promoting cell tolerance to replicative stress (or in suppressing genomic instability), I have a few comments the authors should consider to improve their manuscript.

(1) I am not sure how cells have gained a growth advantage if they were arrested (Line 105-106). Perhaps the authors can elaborate.

Thanks for pointing this out and we are sorry for the misleading statement. We have revised the manuscript and would like to clarify that “survival advantage” may be more accurate than “growth advantage”, and since long-term DOX treatment led to decreased cell survival indicated by decreased number of colonies in Supplemental Fig. S1D, thus many cells died during DOX treatment. Therefore, the cells able to survive throughout DOX treatment and being collected for sequencing may have gained survival advantage compared to their counterparts who fail to survive.

(2) Figure 3D - From Western blotting of ELF3, forced expression of E2F6 does not appear to "block" HU-induced ELF3 up-regulation, but merely down-regulate basal level of ELF3, with the effect of HU still notable.

Thanks for the comment and we agree that E2F6 down-regulate ELF3 baseline expression levels and did not fully block ELF3 up-regulation. After calculating the foldchange after E2F6 overexpression, we did confirm that E2F6 overexpression still partially block HU-induced ELF3 up-regulation, with foldchange from 3.32 to 2.40, supporting our conclusion that HU-induced ELF3 upregulation is regulated by ATRChk1-E2F axis. It does, however, cannot be excluded that E2F6 also regulates ELF3 expression in other replication stress-independent ways, and we have revised the manuscript accordingly.

(3) Figure 3J & K - In my opinion, if BRCA1 knockdown were more efficient it remains formally possible that co-depletion of BRCA1 and GATA3 may exhibit additive effects in up-regulating ELF3 mRNA level.

Thank you for the comment. Actually, the BRCA1 knockdown efficiency in Figure 3J was shown in Supplemental Fig. S3B, and notably both BRCA1 and GATA3 knockdown were numerically more efficient in the double-knockdown group than in the single-knockdown group, individually. Thus, the higher ELF3 up-regulation in double-knockdown group in Figure 3J could be cause by the superior knockdown efficiency of both BRCA1 and GATA3. Nonetheless, we agree that it might be possible that BRCA1 and GATA3 still have separate functions in this experimental setting and marginal additive effect may exist, and the manuscript was revised accordingly.

(4) Figure 4 - Perhaps the authors can change its title to better summarise the findings. Cell sensitivity assays and xenograph experimentations may not necessarily relate to genomic instability.

Thank you for the great suggestion. To summarize the results more accurately, we have revised the title as “ELF3 can help cells tolerate replication stress and sustain cell survival”.

(5) Figure 5B&C - It would be important to document the time-dependent resolution of HU-induced DNA lesions by including additional time-points before, during, and after HU treatment.

We appreciate the suggestion to include additional time points to document the timedependent resolution of HU-induced DNA lesions. In our experiments, we observed that ELF3 knockdown leads to genomic instability both in the presence and absence of HU treatment. Specifically, Figure 5A and Figure S5 demonstrate that ELF3 knockdown increases genomic instability without HU treatment, indicating its role in maintaining genomic stability under normal conditions. On the other hand, Figure 5B, 5C, and 5D show that ELF3 knockdown under HU-induced replication stress further exacerbates genomic instability. This observation aligns with our finding that ELF3 expression increases in response to replication stress, suggesting its critical role in maintaining replication homeostasis under such conditions.

1. Figure 5F&I - Which ELF3 siRNA was used in these experimentations? Since the authors did not exclude off-target effects perhaps it may be worthwhile to include both ELF3 siRNAs for Panel F.

Thanks for your advice. The qPCR (Figure 5F) and DNA fiber assay (Figure 5I) were using siELF3-4 siRNA. And we repeat the qPCR experiments for Panel F using siELF3-5 siRNA (Supplement Fig. S5B).

We sincerely thank you for your thoughtful feedback and constructive suggestions. Addressing these points has strengthened our manuscript, and we are grateful for the opportunity to refine and clarify our work. We appreciate your critical evaluation and look forward to further constructive dialogue.

Reviewer #2 (Recommendations For The Authors):

(1) The data driving the hypothesis uses gene expression signatures as an indirect measure of replication stress. This is a critical concern.

a. At this time, numerous gene expression signatures have been reported to be biomarkers of replication stress. Therefore, it would be valuable to apply additional gene expression signatures to examine the performance and the overlap in the results.

The recent work by Takahashi et al., 2022 (https://pubmed.ncbi.nlm.nih.gov/36381660/) provides a signature that was derived independently and offers one that can be used to assess the performance of the signatures and stability of the conclusions.

Thank you for the valuable suggestion. We have done the replication stress evaluation of mammary cell subgroups using the Repstress score developed in the work you mentioned. The result showed that LP cells have trends of higher replication stress compared with other subgroups, though no statistical significance. This result, consistent with our previous analysis, indicated that LP cells have higher trends of replication stress levels. And we have added this data as the last line of Figure 1A in revised vision.

Replication stress pathway scores of different human normal mammary cell populations.

The gene expression data were from Lim et al. (3).

b. A direct measure of replication stress in LP cells would be important to confirm the gene expression signature. Therefore, performing immunostaining for markers of replication stress (eg gamma-H2AX foci, DNA fiber assays) would provide more direct data to support the assertions.

Thank you for this suggestion and we totally agree that experiments revealing replication stress levels by investigating common markers, e.g., gamma-H2AX foci, DNA fiber assays, will provide vital evidence for our hypothesis. However, since our last response, we have been diligently trying to obtain LP cells for these experiments but encountered technical challenges while attempting to isolate and culture LP cells in vitro.

In the discussion part, we have revised the manuscript to emphasize that the data obtained from MCF10A should be interpreted with caution and there are certain gaps between the cell models and LP cells.

(2) The depth of single-cell sequencing can often be limiting. Therefore, a supplementary table listing the genes used for the replication stress signature and the frequency that they are observed in the single-cell sequencing data. This is needed to ensure that the replication stress score does not reflect a small subset of the replication stress signature genes.

Thanks very much for this evaluable suggestion. We have provided an expression matrix of genes for the replication stress signature in the revised version (Supplementary Table S1), And we also calculated the average expression level of each gene in the cells. As shown in Author response image 2, these genes expressed relatively low at the single-cell level (with counts≤10), The expression differences among genes are relatively small. Thus, we excluded the possibility that several high-expressed genes significantly affect the replicative stress score.

Average counts of Top 50 genes for the replication stress signature

(3) As only 4 BRCA mutation carriers are analyzed, it is critical that the mutations be reported for these individuals because pathogenic variants differ in their effects and interactions with the DNA repair machinery in cells.

Thanks for the suggestion and the information of 4 BRCA1 mutant carriers were added in Supplemental Table S2.

(4) The figures throughout lack critical details making it difficult to evaluate. Figure 1A states that these are "replication stress pathway scores..." but there is no evaluation of levels of statistical differences. The heat map has what appears to be a log unit score between +2 and -2 but it is unclear whether it is log2 or log10 or some other unit. In 1B, the replication stress scores are visualized as relative values between 0 and 0.1, but there is no indication of what this means or whether there is a statistically significant difference in the levels among the populations. As tumors are composed of multiple cell types, it should be stated how the "tumor cells" are uniquely identified in the figure legend. The lack of critical information is common across many of the figures making review frustratingly difficult.

Thanks for the suggestion. We have added the statistical analysis and scale in Figure 1A legend. For Figure 1B, replication stress was calculated by sum of replication stress gene expression and presented as ln value. We have provided a quantitative figure and statistical tests (by Mann-Whitney) of replication stress scores for various cell types (Supplementary Figure 1A).

In addition, we added details of identification of tumor cells in the method section in the revised manuscript. Briefly, the adjacent normal breast sample served as a control to filter various types of normal cells from tumor samples. the normal cells from the tumor sample were merged with the same types of normal cells from adjacent normal breast samples, leaving one cell cluster only generalized by tumor sample. These tumor specific clusters were considered as malignant cell populations. We further found that the malignant cell population showed higher UMI counts than the normal cell populations, consistent with active metabolism in the malignant cells. More importantly, ER, PR, and HER2 expression of the malignant cells in each case were exactly matched with the clinical records. Finally, we utilized InferCNV to validate malignant cells subset as higher copy number alterations (CNAs) detected in the malignant cells compared with normal cells.

(5) The hypothesis states that the LP cells are uniquely sensitive to deficiency in BRCA1 compared to other cells. However, the authors use knockdown of BRCA1 in MCF10A cells which are generally considered to be basal cells and not LP cells.

Thanks for the comment. We totally agree that MCF10A cannot reflect the LP features and was mainly used as a normal mammary cell line model. We have tried to obtain human LP to perform some experiments but have all failed due to the cell vulnerability and difficult to be passed on in vitro. The gap between MCF10A and LP cells was stressed in the discussion part.

(6) Figure 2, the number of samples being compared is not listed for most of the panels. It appears that ELF3 is enriched in subsets of breast cancers, but much of the data is not focused on BRCA1-deficient tumors. Therefore, the data appears to show that ELF3 expression is more of a generalized feature of TNBCs (which has been reported previously) and dilutes the support for the hypothesis. Therefore, panels C-G raise concerns regarding the overall hypothesis that LP cells are the cell type that is affected.

Thanks for the suggestion. We have added the number of samples in Figure 2 legends.

Our analysis focus on basal subtype because of the well-known relationship between BRCA1 deficiency and this subtype. Our results demonstrate the association between ELF3 expression and basal, TNBC, as well as HER2+ subtype, consistent with previous reports. Since TNBC also has high replication stress levels (NPJ Breast Cancer. 2020 Sep 7;6:40.), ELF3 upregulation in this subtype may not be solely due to BRCA1 deficiency, and we totally agree that this analysis may dilute the relationship between ELF3 and BRCA1. We have revised the discussion part to be more precise on this.

(7) Figure 3 provides experimental support for the hypothesis. While panel A is of interest, the legend lacks any description beyond "normal mammary tissue" and that there are non-carriers and carriers of BRCA1 mutations. Is this from bulk RNAseq data or single-cell RNAseq data? How many carriers and how many noncarriers? Panel E is ENCODE data from MCF7 cells that are ER+ luminal subtype so it is unclear if this is relevant to the LP cells that are the focus of the hypothesis.

Thanks for the comments. Figure 3 panel A was from single-cell RNAseq data, including 3 BRCA1 WT patients and 4 BRCA1 mutant patients. All cells (normal cells and tumor cells) are involving, and ELF3 expression was normalized by reads in each cell. We have added this information in the figure legend.

It has been difficult to obtain ENCODE data in LP cells. The effect of E2F1 on regulation of ELF3 was validated in MCF10A cells by experiment and consistent with MCF7 ENCODE data, thus we suggest this effect can be conserve in mammary cells, but further confirmation in LP cells is needed. We have revised the manuscript to note that.

(8) In Figure 4, the authors use BRCA1-deficient breast cancer cells to show the reliance on ELF3 and suggest that this is specific to this genetic lesion and not other subtypes. However, there is no data to show that this is not observed using ER+ cells or TNBC that are not BRCA1-deficient cell lines or models.

Thanks for pointing this out. As ELF3 knockdown in MCF10A resulted in increased genomic instability (Supplement Fig S5) and less capability to resolve replication stress (Figure 5B), we believe that ELF3 can help deal with replication stress not specifically in BRCA1-mutant cells, but also normal mammary cells, and also multiple cell lines with distinct backgrounds as suggested in Figure 4G, 4H and Supplement Fig S4G. The special link between ELF3 and BRCA1 is reflected by ELF3 significant upregulation upon BRCA1 deficiency, but not ELF3 downstream functions.

(9) Figure 5 provides the first direct evaluation of biomarkers of replication stress (gamma H2AX, 53BP1). DNA fiber assays provide the most direct evaluation of replication fork kinetics, and therefore, replication stress. The knockdown of BRCA1 and ELF3 appear to phenocopy one another in the HCC1937, but there is no other cell type to show whether this is specific for BRCA1-deficient cells. For example, the MCF7 cells show E2F1 binding to ELF3 (Figure 3E) and may show replication stress upon knockdown of ELF3. Without testing this, the authors cannot suggest that the effect is linked to BRCA1 status. The authors do not identify the BRCA1 mutation in these cells and whether there is homozygous loss. Similarly, the mutational status in the SUM149PT cells should also be stated. These need to be added to aid interpretation of the results.

Thank you for the constructive advice. We have added information regarding BRCA1 status of HCC1937 and SUM149PT. As discussed before, the results from Figure 4G and 4H suggest that ELF3 expression is associated with sensitivity to replicationstress-inducing-drugs across many cell lines. Thus ELF3 can maintain the stability of DNA replication is not specific to BRCA1-deficient cells. The reliance of ELF3 in BRCA1-deficiency we proposed is mainly focus on the fact that ELF3 is upregulated in BRCA1 deficient conditions, plus ELF3 may help cells tolerate replication stress during the transformation, therefore the resulted tumor cells-that is BRCA1-deficient breast cancer cells-may be more sensitive when losing ELF3 expression.

(10) While the data in Figure 6 are valuable extensions of the gene signature derived from the MCF10A cells with BRCA1 knockdown, only 2 BRCA1 carriers are reported. As carriers bear heterozygous mutations in BRCA1, haplo-insufficiency would be necessary to generate the signature. The authors do note the publication by Panthania et al, but there are relatively few examples of haploinsufficiency. It should be noted that Sedic et al., 2015 also suggested haploinsufficiency in breast epithelial cell cultures from BRCA1 heterozygotes which appears to cause premature senescence, possibly via replication stress. However, this was observed in the basal epithelial cells. Therefore, this appears to be a feature of the breast epithelium more generally and is not enriched or limited to the LP cells.

Thanks very much for your valuable suggestion. We have revised the discussion part to involve this important work and we fully agree that BRCA1 deficiency can cause replication stress not limited to LP cells. While in fact, the point we would like to address in Figure 6 is that BRCA1 deficiency modules the transcription profile towards LP-like cells, but not other-subtype-like cells, in normal mammary cells. We observed surprisingly similar profile between BRCA1-deficient cells and LP cells, suggesting there might be an inherent function of BRCA1 to mediate LP genes transcription. Furthermore, the data indicate that ELF3 has a tighter association with LP genes than other recognized LP-specific transcription factors like ELF5 and EHF, which are of the same family of ELF3. This result is intriguing since ELF3 can be upregulated by BRCA1 deficiency and replication stress. We assume that ELF3 could be a transcription node downstream of BRCA1 deficiency and modulate LP genes expression, and this process might be limited to LP cells since ELF3 has the highest expression levels in LP. Nonetheless, this hypothesis is also needed to be validated in LP cells by experiments.

We would like to express our deepest gratitude to the reviewers for their thorough and constructive feedback. Their insightful comments have been invaluable in guiding the revisions of our manuscript, helping us to clarify our hypotheses and strengthen the presentation of our findings. While we encountered some challenges, particularly with the isolation and culturing of LP cells, we made significant efforts to address the reviewers' concerns to the best of our ability. We have updated our manuscript accordingly, ensuring that all issues raised have been addressed comprehensively. We believe that these revisions have substantially improved the quality and clarity of our work, and we are excited to share our findings with the scientific community. Thank you once again for the opportunity to revise our manuscript, and we look forward to your feedback on the updated version.

Version 2

Abstract

Abstract

BRCA1 is a critical tumor suppressor, mutations in which greatly increase risks for many tumors in carriers, most notably breast cancer. Luminal progenitor cells (LPs) are the currently recognized cells origin of BRCA1-deficient breast cancers. However, the reason why LPs are prone to transform with BRCA1 deficiency has not been elucidated. Here, using single-cell sequencing of human BRCA1 mutant breast cancers and RNA sequencing (RNA-seq) of BRCA1-deficient normal mammary cells, we reveal that replication stress is a feature of LPs and a driving factor during BRCA1-associated tumorigenesis. Mechanistically, replication stress and BRCA1 deficiency lead to significant upregulation of ELF3 expression. ELF3 can help suppress excessive genomic instability and promote LP transformation with BRCA1 deficiency. Moreover, ELF3 emerged as a core transcription factor regulating LP genes, leading to LP expansion. Our findings suggest that replication stress is a driving factor during BRCA1-associated tumorigenesis in luminal progenitor cells and elucidates the key role of ELF3 during this process.

Introduction

BRCA1 is one of the most important tumor suppressor genes. Carriers of BRCA1 germline mutations are at greatly increased risks for developing various tumors, including breast, ovarian, prostate, colon, pancreatic, and gastric cancers (1, 2).Notably, BRCA1 is most closely associated with breast cancer. However, why BRCA1-associated tumors exhibit significant tissue specificity is unknown, and the effects of BRCA1 defects in normal human mammary cells are largely obscure.

In addition to the tissue specificity of tumors caused by BRCA1 deficiency, breast cancers associated with BRCA1 deficiency exhibit a set of other distinctive features. Most BRCA1 mutation carriers develop triple-negative breast cancers (TNBCs), in which tumors show negative expression of estrogen receptor (ER), progesterone receptor (PR), and Her2 (human epidermal growth factor receptor 2), and these cancers are usually also classified as basal-like breast cancers. The currently recognized cells of origin of BRCA1- deficient breast cancers are the luminal progenitor cell (LPs) in the mammary gland (3–6). LPs show abnormal expansion in the normal mammary tissue of BRCA1 mutation carriers (3, 7), which might be caused by a differentiation blockade (8). Investigating the traits of LPs is important for tumor prevention and treatment in BRCA1 mutation carriers and a wider range of patients with TNBCs and basal-like breast cancer. However, the reason LPs are prone to transform into cancer cells when BRCA1 is deficient is unclear.

ELF3 is a member of the ETS family of transcription factors. Early reports suggested that ELF3 may regulate the development and differentiation of mammary tissue (9). Intriguingly, ELF3 has opposite functions in tumorigenesis in different tissues. In breast cancers, ELF3 shows high expression (10), plays an oncogenic role by promoting EMT (11) and forms a positive feedback loop with Her2 to maintain the transformation phenotype of Her2+ breast cancers (12). However, the mechanism of high ELF3 expression in certain types of breast cancer and its association with breast cancers beyond the Her2+ subtype is currently unknown.

In this study, utilizing single-cell sequencing of BRCA1 mutant breast cancers and time-resolved RNA- seq of BRCA1-deficient normal mammary cells, we revealed that LPs have a trait of higher replication stress compared to other cell populations in the normal mammary tissue, endowing them with the potential for transformation when BRCA1 is deficient, and ELF3 plays a key role in the process. Mechanistically, ELF3 expression is significantly upregulated by replication stress and BRCA1 deficiency, dependent on the ATR-Chk1-E2F axis and the transcriptional regulation function of BRCA1, respectively. In addition, ELF3 can help suppress excessive genomic instability, promoting the tolerance of LPs to BRCA1 deficiency. Notably, BRCA1 deficiency causes cells to display LP transcriptional profile characteristics, leading to dedifferentiation and expansion of LPs, and ELF3 is a core transcription factor during this process. Our study reveals why BRCA1 deficiency is prone to result in tumorigenesis in LPs and elucidates the key role of replication stress and ELF3 during this process, suggesting that ELF3 could be a promising target for BRCA1-associated breast cancers.

Results

LPs have higher levels of replication stress compared with other mammary cell populations

LPs in the mammary gland are the currently recognized cells of origin of BRCA1-deficient breast cancers. However, the reason for LPs are prone to transform with BRCA1 deficiency has not been elucidated. Firstly, we considered that LPs, which are stimulated by upstream progesterone paracrine signals through RANKL (5), may have higher levels of replication stress, which can trigger tumorigenesis by increasing genomic instability. We utilized both published data and our published single-cell sequencing data (8) to address this question. We evaluated the replication stress levels in different mammary cell populations from the dataset of Lim et al. (3) by scoring multiple DNA replication pathway gene expression as previously studied (13, 14). The result showed that LPs have the highest level of replication stress among all subgroups of normal human mammary cells (Fig. 1A, Supplemental Table S1). We further analyze our previously obtained single-cell sequencing data of four BRCA1 mutation carriers (case #1-4) and normal mammary cells of three matched non-carriers (case #5-7). According to our previous analysis, case #3 BRCA1 mutant carrier developed basal-like breast cancer, which was derived from LPs. Importantly, in this case of BRCA1 mutant carrier, single-cell sequencing data revealed that tumor cells had the highest levels of replication stress, and LPs had the second-highest levels (Fig. 1B, Supplemental Fig. S1A and Supplemental Table S2). Moreover, we applied Monocle to reconstruct the transformation trajectories from normal mammary cells to tumor cells in the BRCA1 mutation carrier and found that the levels of replication stress showed an increasing trend during the process of LP transformation into tumor cells (Fig. 1C and Supplemental Fig. S1B). These data demonstrate that replication stress is an important trait of LPs and might be a key driver of BRCA1-deficient tumorigenesis, which is consistent with the previous report that BRCA1 mutant mammary epithelial cells are defective in replication stress suppression (15).

LPs have higher levels of replication stress compared with other mammary cell populations.

(A) Replication stress pathway scores of different human normal mammary cell populations, scaled by row, transformed to mean=0 and variance=1. The gene expression data were from Lim et al. (3). The gene expression signatures were from Dreyer et al. (13) and Takahashi et al. (14).

(B) T-distributed statistical neighbor embedding (t-SNE) plots visualizing the replication stress scores and ELF3 expression levels in breast cancer and normal mammary tissue from a BRCA1-mutated triple- negative breast cancer patient. The cells are colored by cell type (top) and replication stress score (bottom).

(C) Monocle-generated cell trajectories visualizing the change in replication stress scores during tumorigenesis. The tumor cells and compared normal luminal cells were obtained from a triple-negative breast cancer patient (case #3) carrying a BRCA1 germline mutation. LP, luminal progenitor cells; ML, mature luminal cells.

(D) Volcano plots showing DEGs in MCF10A-shBRCA1-Tet-on cell RNA-seq results after 1 μg/mL DOX treatment for 2 (left), 5 (middle) and 10 (right) days versus the Ctrl group. DEGs are genes with |log2-fold change| > 2 and FDR < 0.001.

(E) PCA plot of MCF10A-shBRCA1-Tet-on cell RNA-seq results.

(F) Mfuzz clustering analysis of Day 10 versus Ctrl DEGs. The heatmap (left) shows the differential expression mode of all genes in each cluster. The line chart (right) shows the global changing mode of each cluster.

(G) Overlap of HU treated MCF10A RNA-seq results and MCF10A-shBRCA1-Tet-on cell RNA-seq Cluster1 genes.

Next, we aimed to find other key regulators during BRCA1-deficient tumorigenesis in LPs. In addition to inducing replication stress, BRCA1 deficiency also profoundly affects the transcriptional profile of normal mammary cells by disrupting the transcriptional function of BRCA1, which may further boost the process of tumorigenesis. To investigate the transcriptional impact of BRCA1 deficiency in normal human mammary epithelial cells, we took advantage of the Tet-on system to build an inducible knockdown model in MCF10A cells. By adjusting the treatment condition of DOX (doxycycline), different time points of BRCA1 deficiency can be controlled. Treatment with DOX causes a constant and efficient knockdown of BRCA1, verified at mRNA and protein levels (Supplemental Fig. S1C). We showed that DOX treatment induces significant growth arrest and decreased cell survival (Supplemental Fig. S1D) as well as an increase in genomic instability (Supplemental Fig. S1E), consistent with the phenotypes of BRCA1- deficient cells (16), confirming that our system functioned as expected. We then performed time-resolved RNA-seq with different durations of DOX treatment (2, 5 and 10 days; Supplemental Fig. S1F). Notably, long-term DOX treatment resulted in decreased number of colony formation, indicating impacted cell survival, thus cells collected for sequencing at the end of 10-days DOX treatment may have gained survival advantage compared to cells that did not survive throughout the DOX treatment. Differential gene expression analysis revealed that BRCA1 was efficiently knocked down during the whole experiment (Fig. 1D). Importantly, the principal component analysis (PCA) revealed that samples showed a time-course distribution in the dimension of PC1, which has the dominant power of explanation (Fig. 1E and Supplemental Fig. S1G). These data demonstrate that the variations between samples can be largely attributed to time. To further investigate the modes of differentially expressed genes (DEGs) that change over time, we carried out the DEG cluster analysis using Mfuzz (17). Analysis of the Day 10 DEGs showed that genes could be divided into 5 clusters with different expression change patterns (Fig. 1F). We mainly focused on Cluster 1, for these sharply upregulated genes may play more important roles in cells gaining survival advantage with BRCA1 deficiency at Day 10, mimicking the initiation of BRCA1 deficiency associated tumorigenesis. To find out the most potent regulator upregulated both by replication stress and BRCA1 deficiency, we performed RNA-seq in HU-treated MCF10A cells and overlapped the upregulated genes in HU RNA-seq and Cluster1 genes in BRCA1 deficiency RNA-seq (Fig. 1G). The results showed that ELF3, a transcription factor, is one of the most dominant genes in this analysis. ELF3 has been reported to be involved in functions including EMT in human breast cancer and promotes the transformation of normal human mammary cells (11, 18). Therefore, we speculated that ELF3 could be an important driving factor during the tumorigenesis of BRCA1-deficient and basal-like breast cancer.

ELF3 is upregulated in BRCA1-associated breast cancer and is related to a worse prognosis

We next explored the expression of ELF3 in breast cancers in the TCGA and METABRIC databases to further investigate the association between ELF3 and BRCA1 and validate our previous analysis. In both clinical databases, ELF3 expression is significantly higher in BRCA1-associated breast tumors than in non-BRCA1-associated breast tumors (Fig. 2A). Moreover, the expression levels of BRCA1 and ELF3 showed a significant negative correlation, consistent with our findings that ELF3 was upregulated under BRCA1 deficiency (Fig. 2B). Since most BRCA1 mutation carriers develop basal-like breast cancers, we next investigated the association between ELF3 and basal-like breast cancers. We exploited the TCGA and METABRIC databases to compare ELF3 expression levels among different breast cancer subtypes. As we expected, ELF3 showed the highest expression levels in basal-like breast cancer in both databases (Fig. 2C). We took advantage of novel subgroups identified by the METABRIC (19) and further explored ELF3 expression in ten IntClusts of the METABRIC database. Among all subgroups, IntClust 5 was enriched with Her2+ cancers, and IntClust 10 was enriched with basal-like cancers. These two subgroups showed the highest expression levels of ELF3, with no significant difference between them (Fig. 2D). The higher levels of ELF3 expression in Her2 subtype breast cancer are consistent with previous studies showing that ELF3 may be a downstream transcription target of Her2 and that there might be a positive feedback loop between them (12, 20, 21). On the other hand, IntClust 4 shows a favorable outcome and has a ‘CNA- devoid’ feature, which has the lowest level of ELF3 expression. These data are consistent with our findings that ELF3 is upregulated by BRCA1 loss, indicate that ELF3 may play an important role in basal-like breast tumorigenesis, and moreover, the expression of ELF3 may be related to genomic stability status. Furthermore, among human breast cancer tissue samples, ELF3 expression levels were significantly higher in TNBCs than in other subtypes (Fig. 2E and F). Additionally, TNBCs have higher ELF3 expression levels than the Her2 subtype (Fig. 2E and F), revealing a tighter connection between ELF3 and BRCA1-associated breast cancers.

ELF3 is upregulated in BRCA1-associated breast cancers and is related to a worse prognosis.

(A) ELF3 expression levels in BRCA1-associated breast cancers and BRCA1-nonassociated breast cancers in the TCGA and METABRIC databases. BRCA1-associated breast cancers comprise samples with BRCA1 mutations, heterozygous loss and homozygous deletion (Mann–Whitney U test)(TCGA: BRCA1-associated, N=362; BRCA1-nonassociated, N=720; METABRIC: BRCA1-associated, N=540, BRCA1-nonassociated, N=1364).

(B) Correlation of ELF3 and BRCA1 expression levels in TCGA breast cancer datasets.

(C) ELF3 expression levels of different molecular subtypes of breast cancers in the TCGA and METABRIC databases (one-way ANOVA).

(D) ELF3 expression levels of different IntClusts of breast cancers in the METABRIC database (one-way ANOVA; for IntClust 5 and 10, Mann–Whitney U test, ns, no significance).

(E) Immunohistochemistry images of human triple-negative, Her2+ and ER+ PR+ Her2+ breast cancer samples stained with the ELF3 antibody.

(F) ELF3 expression levels of the indicated types of human breast cancers (Mann-Whitney U test, *, p < 0.05).

(G) Survival analysis of the indicated subtypes of human breast cancer in the KMplot database. For the TNBC group, samples with negative expression of all hormone receptors (ER, PR and Her2) are included. In the TNBC group, high expression (n=54), low expression (n=62); in the basal group, high expression (n=85), low expression (n=211); in the Lum A group, high expression (n=121), low expression (n=101); in the Lum B group, high expression (n=133), low expression (n=67); and in the Her2 group, high expression (n=72), low expression (n=126).

ELF3 shows significant copy number amplification in breast cancer (22). Thus, we wondered that whether the high expression levels of ELF3 in BRCA1-associated breast cancers are due to copy number variation. Therefore, we analyzed the proportion of ELF3 copy number amplification in different subtypes (Supplement Fig. S2) and found that this was not the case. In TCGA basal-like breast cancer, the ELF3 amplification proportion was not higher than that in other subtypes but was only 3.5%, which was lower than the total level (7.3%). These data demonstrate that the upregulation of ELF3 in BRCA1-associated breast cancer is not caused by copy number amplification.

Furthermore, we investigated whether ELF3 expression could be a biomarker for breast cancer prognosis. We found that higher ELF3 expression is correlated with worse prognosis in TNBC and basal-like breast cancer (Fig. 2G). In the Her2 subtype, higher ELF3 expression is also related to shorter overall survival, consistent with previous reports that as a Her2 target gene, ELF3 may help promote tumor invasion (21). Interestingly, ELF3 expression shows less or no prognostic prediction value in luminal A and luminal B subtypes, suggesting that ELF3 plays a more powerful role in TNBC and basal-like breast cancer. These data imply that in BRCA1-associated breast cancer, ELF3 tends to be upregulated and plays a “driver” role, promoting tumor progression and leading to a worse prognosis.

ELF3 upregulation is induced by replication stress via the ATR-Chk1-E2F axis and by BRCA1 deficiency via GATA3 transcription

We next focused on the mechanisms of ELF3 upregulation in the presence of replication stress and BRCA1 deficiency. Consistent with the HU RNA-seq analysis, our single-cell RNA-seq data showed that in both BRCA1 WT and mutant carriers, the levels of replication stress and ELF3 expression showed a significant positive correlation (Fig. 3A). ELF3 was also upregulated by HU-induced replication stress in MCF10A cells (Fig 3B). Since ATR is the key kinase that responds to replication stress, we treated cells with ATRi and found that it could efficiently block ELF3 upregulation induced by HU, while ATMi had no such effect (Fig. 3B). Moreover, downstream of ATR, ELF3 expression upregulation depended on Chk1 activation but not Chk2 activation (Fig. 3C). It has been reported that the transcriptional activation induced by replication stress-ATR-Chk1 pathway largely depends on E2F transcription factor activation (23). Under replication stress, ATR-Chk1-mediated phosphorylation of E2F6 blocks its inhibitory function on E2F and activates E2F transcription. Thus, we overexpressed E2F6 and found that E2F6 largely blocks the upregulation of ELF3 under HU treatment conditions (Fig. 3D), validating that ELF3 upregulation relies on E2F6 inactivation and E2F de-suppression. Moreover, utilizing ENCODE ChIP-seq data, we found that the ELF3 promoter is indeed bound by E2F1 (Fig. 3E). These data demonstrate that ELF3 is upregulated by the ATR-Chk1-E2F axis under replication stress conditions.

ELF3 upregulation is induced by replication stress via the ATR-Chk1-E2F axis and by BRCA1 deficiency via GATA3 transcription.

(A) Spearman correlation analysis of ELF3 expression levels and replication stress scores in normal mammary tissue from noncarriers (left) and BRCA1 mutation carriers (right). Expression data is from single-cell RNAseq data, including 3 BRCA1 WT patients and 4 BRCA1 mutant patients. All cells (normal cells and tumor cells) are involving, and ELF3 expression was normalized by reads in each cell.

(B) Western blot and RT–qPCR results of MCF10A cells treated with the indicated drugs for 12 h. Drug concentration: HU 1 mM, ATRi (VE821) 10 μM, ATMi (KU5593) 10 μM.

(C) Western blot and RT–qPCR results of MCF10A cells treated with the indicated drugs for 12 h. Drug concentrations: HU 1 mM, ATRi (VE821) 10 μM, Chk1i (GDC-0575) 50 μM, Chk2i (BML-277) 10 μM.

(D) Western blot results of MCF10A cells transfected with the empty vector or Myc-E2F6 and treated with DMSO or 1 mM HU for 12 h.

(E) ChIP-Seq enrichment signal (green plot, ENCODE ENCFF858GLM; data are representative of two independent experiments) and the location of detected peaks (black line, ENCODE ENCFF692OYJ; irreproducible discovery rate (IDR) thresholded peaks) of E2F1 in MCF7 cells from the ENCODE database.

(F) ELF3 expression changes as BRCA1 deficiency duration increases. ELF3 expression levels are indicated by the CPM of the RNA-Seq results (means ± SD, n=3).

(G) Western blot and RT–qPCR results of MCF10A cells transfected with siNC or siBRCA1 for 48 h.

(H) RT–qPCR results of MCF10A cells transfected with siNC or siGATA3 for 48 h.

(I) ChIP-Seq enrichment signal (yellow plot up, ENCODE ENCFF384CPN; yellow plot down, ENCODE ENCFF342GNN. Data are representative of two independent experiments) and the location of detected peaks (black line up, ENCODE ENCFF352QVM; black line down, ENCODE ENCFF437NQS; both are irreproducible discovery rate (IDR) thresholded peaks) of GATA3 in MCF7 cells from two independent studies (ENCSR423RTK and ENCSR000BST) from the ENCODE database. The motif sequences were obtained from the JASPAR database.

(J) RT–qPCR results of MCF10A cells transfected with the indicated siRNA for 48 h.

(K) Western blot and RT–qPCR results of MCF10A cells transfected with siNC or siBRCA1 for 24 h followed by treatment with DMSO or 1 mM HU for 12 h.

(L) RT–qPCR results of MCF10A cells transfected with siNC or siGATA3 for 24 h followed by treatment with 1 mM HU for 12 h. In all panels of RT–qPCR results, data are presented as the mean ± SD, n=3, *, p < 0.05; **, p < 0.01; ***, p < 0.001; ns, no significance, by paired two-tailed Student’s t test.

Then we investigate the mechanism of ELF3 upregulation induced by BRCA1 deficiency. ELF3 expression gradually increased during 10 days of DOX treatment (Fig. 3F). ELF3 upregulation was further confirmed by siRNA-mediated BRCA1 knockdown in MCF10A cells to strictly exclude the unknown influence of the Tet-on system (Fig. 3G). BRCA1 has transcription regulation functions by interacting with other transcription factors to bind to gene promoters and regulate gene expression (24). Therefore, we explored transcription factors associated with BRCA1, including GATA3 (25) and Myc (26), for whether the binding motifs of these transcription factors appear in the ELF3 promoter sequence utilizing the JASPAR database (Supplement Fig. S3A). We found that E2F1 was present in the prediction list with a high relative score, consistent with our results that E2F1 can bind ELF3 promoter. The prediction results show that the motif of GATA3 appeared in the ELF3 promoter multiple times with an overall high relative score. Indeed, knockdown of GATA3 in MCF10A cells also strongly induced ELF3 expression (Fig. 3H). Similarly, we found multiple GATA3 binding peaks at the promoter of ELF3 in two independent ChIP- seq datasets from the ENCODE database, consistent with two of the GATA3 binding sequences predicted in the JASPAR database (Fig. 3I). These data suggest that ELF3 is a GATA3 downstream target gene. Importantly, when both BRCA1 and GATA3 were knocked down, there was no significant additive effect on ELF3 upregulation (Fig. 3J and Supplement Fig. S3B), implying that BRCA1 and GATA3 mainly act in the same pathway for ELF3 induction.

Notably, we found that and BRCA1 knockdown and HU treatment have a superimposed effect on upregulating ELF3 expression levels (Fig. 3K and Supplement Fig. S3C). Furthermore, GATA3 knockdown and HU treatment also had a significant superimposed effect in upregulating ELF3 expression (Fig. 3L and Supplement Fig. S3D). These data validate that the BRCA1-GATA3 axis and replication stress regulates ELF3 expression independently.

ELF3 helps cells tolerate replication stress and sustain cell survival

Next, we investigated the impact of ELF3 upregulation under replication stress conditions and BRCA1 deficiency. Based on the association between ELF3 and the worse prognosis in BRCA1-associated breast cancer patients (Fig. 2G), we hypothesized that upregulation of ELF3 could help cells deal with replication stress and suppress excessive genomic instability, facilitating cancer initiation and cancer evolution. As expected, ELF3 knockdown greatly increased the sensitivity of MCF10A cells to HU and cisplatin (Fig. 4A and B, Supplement Fig. S4A). This finding suggests that ELF3 can help normal cells deal with replication stress. Furthermore, ELF3 knockdown in HCC1937 and SUM149PT cells, both of which are human breast cancer cell lines with BRCA1 mutations and allelic loss, also caused a significant decline in cell proliferation (Fig. 4C and D, Supplement Fig. S4B and C), suggesting that ELF3 is essential for BRCA1 mutant breast cancer cell survival. Moreover, HCC1937 and SUM149PT cells became more sensitive to HU and cisplatin when ELF3 was deficient (Fig. 4E and F, Supplemental Fig. S4D and E). Since PARP inhibitors can induce replication fork stalling by PARP1 trapping and lead to replication stress in BRCA1-deficient cells, we also tested olaparib sensitivity in HCC1937 cells and obtained similar results (Supplement Fig. S4F). These results demonstrate that in BRCA1 mutant breast cancer cells, ELF3 can help cells tolerate replication stress and sustain cell survival.

ELF3 helps suppress excessive genomic instability.

(A and B) Cell viability curve of MCF10A cells transfected with siNC or siELF3 for 48 h, treated with the indicated drugs and measured by the CCK-8 assay (mean ± SD, n=3, two-way ANOVA, **, p < 0.01; ***, p < 0.001; ****, p < 0.0001).

(C and D) Cell viability of HCC1937 (C) and SUM149PT (D) cells transfected with siNC or siELF3 for 5 days detected by the CCK-8 assay (mean ± SD, n=4, paired two-tailed Student’s t test, **, p < 0.01; ***, p < 0.001; ****, p < 0.0001).

(E and F) Cell viability curve of HCC1937 cells transfected with siNC or siELF3 for 48 h, treated with the indicated drugs and measured by the CCK-8 assay (mean ± SD, n=3, two-way ANOVA, **, p < 0.01; ***, p < 0.001; ****, p < 0.0001).

(G and H) Correlation of ELF3 expression levels and cell sensitivity of the indicated drugs in the CellMiner (G) and GDSC (H) databases.

(I) Nude mouse tumors of SUM149PT cells infected with shCtrl of shELF3 lentivirus (mean ± SD, unpaired two-tailed Student’s t test, *, p < 0.05).

In addition, utilizing the CellMiner database, we found that among nearly sixty diverse human cancer cell lines, lower ELF3 expression is correlated with higher drug sensitivity to cisplatin and etoposide (Fig. 4G). Consistent results were obtained using the GDSC database (Fig. 4H, Supplement Fig. 4G). Moreover, in nude mice, SUM149PT tumors with ELF3 deficiency were significantly smaller than those in the control group (Fig. 4I). These data indicate that ELF3 may help normal cells tolerate replication stress and subsequent DNA damage, allowing transformation into malignant tumors. During cancer progression, BRCA1 mutant cancer cells gradually become reliant on high ELF3 expression to preserve a certain degree of genomic stability, making ELF3 a promising therapeutic target and biomarker in BRCA1-associated breast cancers.

ELF3 helps maintain the stability of DNA replication

We then investigated the mechanism by which ELF3 helps tolerate replication stress. ELF3 deficiency leads to more endogenous DNA damage in MCF10A cells (Supplement Fig. S5A). In BRCA1-deficient breast cancer cell line HCC1937, knockdown of ELF3 resulted in a significant increase in spontaneous γH2AX foci (Fig. 5A), demonstrating that ELF3 deficiency leads to more endogenous DNA damage in BRCA1-deficient breast cancer cells. Moreover, with HU treatment and release for repair, ELF3 deficiency resulted in a significant increase in the number of γH2AX and 53BP1 foci per cell and the proportion of cells with damage foci (Fig. 5B-D). Under replication stress conditions, γH2AX and 53BP1 foci represent DNA double-strand breaks formed by collapsed replication forks (27). These data suggest that ELF3 deficiency results in a defective ability to repair stalled replication forks; thus, more DNA damage is generated. The upregulation of ELF3 expression during replication stress helps cells better manage stalled replication forks and prevents excessive DNA damage.

ELF3 helps maintain the stability of DNA replication.

(A) Spontaneous γH2AX foci of HCC1937 cells transfected with siNC or siELF3 for 72 h (mean ± SEM, n > 100, unpaired two-tailed Student’s t test, **, p < 0.01; ****, p < 0.0001).

(B and C) γH2AX and 53BP1 foci of MCF10A cells transfected with siNC or siELF3 for 48 h and treated with 2 mM HU for 4 h followed by release for 20 h (mean ± SEM, n > 100, unpaired two-tailed Student’s t test, *, p < 0.05; **, p < 0.01; ***, p < 0.001).

(D) Proportions of cells with γH2AX (≥ 3) and 53BP1 foci from (B and C) (mean ± SD, unpaired two- tailed Student’s t test, *, p < 0.05; **, p < 0.01; ***, p < 0.001).

(E) Enrichment plot of DNA replication pathways in MCF10A ELF3 knockdown RNA seq.

(F) RT–qPCR results of MCF10A cells transfected with siNC or siELF3 (#4) for 72 h (mean ± SD, n=3, paired two-tailed Student’s t test, *, p < 0.05; **, p < 0.01; ns, no significance).

(G) Replication fork velocity indicated by DNA fiber assay of MCF10A cells transfected with the indicated siRNA for 72 h (mean ± SD, n > 80, unpaired two-tailed Student’s t test, ****, p < 0.0001).

(H) Replication fork symmetry indicated by the DNA fiber assay of MCF10A cells transfected with the indicated siRNA for 72 h (mean ± SD, Mann–Whitney test, ***, p < 0.001; ****, p < 0.0001).

(I) Replication fork stability of MCF10A cells transfected with the indicated siRNA for 72 h (mean ± SD, n > 100, unpaired two-tailed Student’s t test, **, p < 0.01; ****, p < 0.0001).

To further explore how ELF3 contributes to replication stress management, RNA-seq on ELF3- knockdown MCF10A cells was performed. Gene set enrichment analysis (GSEA) showed that multiple pathways associated with DNA replication were significantly downregulated in ELF3-deficient cells (Fig. 5E), which was further verified by RT–qPCR (Fig. 5F and Supplement Fig. S5B). These data indicate that DNA replication pathways are affected by ELF3 deficiency.

DNA fiber assay was performed to more closely investigate the effects of ELF3 deficiency on DNA replication. ELF3 deficiency significantly slowed down the DNA replication progression (Fig. 5G), with significantly reduced symmetry of replication forks (Fig. 5H), suggesting that more replication forks were stalled or unstable. Moreover, with HU treatment, ELF3 knockdown undermined the stability of stalled replication forks (Fig. 5I). Notably, double-knockdown of ELF3 and BRCA1 led to a further decrease in replication fork stability (Fig. 5I). These data reveal that ELF3 expression upregulation when BRCA1 is deficient can help cells deal with replication stress, while simultaneous ELF3 deficiency leads to intolerable levels of genomic stability.

ELF3 is a core transcription factor inducing LP genes expression under BRCA1 deficiency

Finally, we explored the possible reasons for the tissue specificity of BRCA1-associated tumors. It has been reported that BRCA1 deficiency disturbs the development and differentiation of mammary tissue and causes abnormal expansion of the LP population (3, 8, 28). We hypothesized that BRCA1 deficiency might affect the normal differentiation process of LPs by disturbing the transcriptional profile, thus leading to abnormal expansion and transformation of LPs. We analyzed the time-course RNA-seq data using a published signature gene set from LPs (3) and found that over time with BRCA1 deficiency, normal mammary cells gradually show a higher signature score of LP gene expression (Fig. 6A). Similarly, using the sets of genes up-and downregulated in the LP population compared to in other mammary cell populations obtained from the data of Pellacani et al. (29), we found that genes upregulated in LPs also tended to be upregulated in BRCA1-deficient conditions, whereas genes downregulated in LPs tended to be downregulated (Fig. 6B). These results suggest that BRCA1 may broadly affect the transcription profile of mammary cells, and therefore, when BRCA1 is deficient, LPs may tend to maintain a dedifferentiated state, leading to the abnormal expansion and tumorigenic properties of these cells in BRCA1 mutant carriers.

BRCA1 deficiency disturbs the differentiation of LPs.

(A) LP signature score of samples of MCF10A-shBRCA1-Tet-on RNA-seq. The signature score was generated using data from Lim et al. (3).

(B) Barcode plot of LP gene set enrichment in Day 10 vs. Ctrl. LP gene sets were obtained from data from Pellacani et al. (33).

(C) GSEA plots of the transcriptional profile of MCF10A cells with BRCA1 deficiency, MCF10A cells with ELF3 overexpression and normal human LPs. The LP gene set was obtained from the data of Pellacani et al. (33).

(D) Violin plots displaying the distribution of ELF3 expression levels in distinct cell types in normal mammary tissue from noncarriers (left) and BRCA1 mutation carrier (right).

(E) ETS subfamily transcription factor motif enrichment results of LP gene promoters.

(F) Expression levels of ETS subfamily transcription factors in LPs from BRCA1 mutation carrier and noncarrier of the single-cell sequencing data.

(G) ETS subfamily transcription factor enrichment of ATAC-Seq of LPs of BRCA1 deficient mice and WT mice from Bach et al. (37).

Multiple ETS family transcription factors were shown to be involved in human mammary development and differentiation in previous reports. However, the function of ELF3 in LPs has not yet been elucidated. We used GSEA to explore the transcriptional profiles of BRCA1-deficient mammary cells, ELF3- overexpressing mammary cells and human LPs from previous reports (29). Surprisingly, these three transcriptional profiles have significant correlations (Fig. 6C), indicating that ELF3 is not only a core transcription node under BRCA1 deficiency conditions but also, more importantly, that ELF3 plays significant roles in the transcriptional regulation of human LPs. In addition, in our single-cell sequencing data, the expression of ELF3 in LPs is the highest among all mammary cell populations (Fig. 6D and Supplement Fig. S6A). Same results were obtained by analyzing the published data from Pellacani et al. (29) (Supplement Fig. S6B). To investigate ELF3 functions in regulating LP genes transcription, we used SEA (simple enrichment analysis) of MEME to perform motif enrichment of ESE and ELF subfamilies transcription factors in promoters of LP genes, as these transcription factors are more closely related to ELF3 in the ETS family, and several, such as ELF5 and EHF, have been reported to regulate LPs (29–31). ELF3 was at the top of the list (Fig. 6E), demonstrating that ELF3 has higher affinity for the promoters of LP genes. Moreover, our single-cell data revealed that ELF3 expression levels were the highest in LPs compared to other members of the ESE and ELF subfamilies (Fig. 6F), including ELF5, which has been reported to play vital roles in LP differentiation. In addition, we analyzed the ATAC-seq data in Brca1- deficient mouse LPs from Bach et al. (32). The ATAC-seq compared the chromatin accessibility of LPs in Brca1/p53-deficient mice with WT mice, and motif enrichment analysis of transcription factors was performed in regions with increased accessibility of Brca1-deficient mouse LPs. ELF3 was among the top genes (Supplement Fig. S6C) and is significantly enriched among all ETS transcription factors (Fig. 6G). These data further confirm our hypothesis that ELF3 plays an important role in the transcriptional regulation of LPs.

Discussion

In this study, we revealed that LPs have the trait of higher replication stress, an important factor for their potential to transformation. The difference of LPs from other cell populations in normal mammary tissue has received considerable attention in the field. Research in mouse mammary tissue has revealed that the overall luminal epithelial cell population is enriched in mitotic, cell cycle and DNA repair-related pathways under estrogen-and progestogen-stimulated conditions relative to the basal and stromal cell populations (33). Subsequently, Casey et al. (34) found that LPs of mouse mammary tissue are enriched in DNA repair and cell division pathways. A recent preprint from the same group (35) continued the analysis of mouse tissue sequencing data and found that in LPs, DNA repair and replication pathways are significantly upregulated, indicating that LPs are more capable of DNA damage repair and thus more tolerant to BRCA1 deficiency. Furthermore, Nolan E et al. (5) found that RANK+ LPs are highly proliferative and prone to DNA damage. These studies did not reveal the reason for LP tendency to transform. In our study, we focus on the characteristic of LPs compared to other cell populations and using sequencing data from normal human mammary tissues (3), we reveal that the DNA replication pathways and the cellular response to replication stress pathways are significantly upregulated in LPs compared to those in other cell populations. We argue that high levels of replication stress may be the major source of DNA damage in LPs, resulting in DDR gene upregulation in these cells, and the high proliferation feature of LP could be the main reason for the high replication stress.

Replication stress is a hallmark of cancer, fueling cancer evolution by increasing genomic instability (36). LPs have inherently higher replication stress levels, and when BRCA1 is deficient, replication stress is further elevated during the transformation from normal LPs to tumor cells, suggesting that replication stress is an important driver of BRCA1-deficient tumorigenesis and the suppression of replication stress could be a new approach for preventive intervention in BRCA1 mutation carriers with the advantages of being less damaging than prophylactic resection. It was reported haploinsufficiency in breast epithelial cell cultures from BRCA1 heterozygotes, which could lead to premature failure through replication stress(37). Moreover, we found that both replication stress and BRCA1 deficiency itself can induce high expression of ELF3, which also plays an important role in the tendency of LPs to transform with BRCA1 deficiency. First, ELF3 can help suppress excessive genomic instability, which is crucial for cell tolerance and survival with BRCA1 deficiency. We speculate that high ELF3 expression can help LPs maintain moderate levels of genomic instability that allow cells to proliferate and is sufficient to fuel cancer evolution, including subsequent events such as p53 mutation and BRCA1 loss of heterozygosity (LOH). Second, ELF3 upregulation can boost LP gene transcription, leading to LP dedifferentiation and expansion. Finally, as an EMT promoter in mammary tissue, ELF3 can accelerate the transformation of LPs to malignant cells. Therefore, high levels of replication stress and ELF3 expression are important factors resulting in the high tendency for LPs to transform (Fig. 7).

Replication stress inducing ELF3 upregulation promotes BRCA1-deficient breast tumori-genesis. LPs have higher replication stress levels under normal conditions (left) and thus are more likely to exceed the tumorigenesis threshold under BRCA1 deficiency (right). Replication stress and BRCA1 deficiency results in further ELF3 upregulation (top right). ELF3 can suppress excessive replication stress and help LPs maintain moderate levels of genomic instability, which is tolerable for cells to proliferate and enough to fuel cancer evolution (a). In non-LP mammary cells, which exhibit inherently low replicative stress, the ELF3 upregulation may enable cells to endure the increased replicative stress caused by BRCA1 deficiency without leading to malignancy. However, in LP cells, which naturally experience higher levels of replicative stress, this ELF3-mediated mech- anism may increase susceptibility to cancerous transformation. In addition, ELF3 upregulation can boost LP gene transcription, leading to LP dedifferentiation and expansion (b). Finally, as an EMT promoter in the mammary tissue, ELF3 can accelerate the transformation process from LPs to malignant cells (c).

A key feature of BRCA1-associated tumors is the tissue specificity, the mechanism of which has long been a research focus. The transcriptional regulation functions of BRCA1 are critical for the development and differentiation of normal mammary tissue. LPs with abnormal expansion have been observed in BRCA1 mutation carriers. Several important differentiation pathways, such as WNT, Notch and Hedgehog, were downregulated in these abnormal LPs (8), suggesting that BRCA1 deficiency may lead to a differentiation blockade in LPs and, therefore, abnormal expansion. Our RNA-seq data revealed that the transcriptional profile of BRCA1-deficient MCF10A cells has a striking similarity to that of LPs in normal human mammary tissue, and ELF3 might be a core mediating transcription factor. This finding suggests that BRCA1 has a broad regulatory effect on gene expression in normal mammary cells and has specific functions in LPs; thus, its deficiency causes LPs to maintain a dedifferentiated state, leading to abnormal expansion of this cell population, and ELF3 may be a core transcription factor inducing these changes. In addition, our study reveals another layer of the mechanism underlying the tissue specificity of BRCA1- associated tumors—the capability of BRCA1 to regulate ELF3 expression. ELF3 is an epithelial-specific transcription factor, but the functions of ELF3 in different epithelial tissues vary. For example, ELF3 promotes tumor development in breast, prostate, and lung cancers (10, 38, 39), whereas it suppresses tumorigenesis in bladder and oral squamous cancers (40, 41). In addition, ELF3 itself plays an important role in the development and differentiation of normal mammary tissue; thus, dysregulation of ELF3 may disturb these processes. Therefore, we suggest that replication stress caused by BRCA1 deficiency leads to upregulation of ELF3 expression in LPs, thus promoting the tissue-specific oncogenic functions of ELF3 and breast tumorigenesis.

This study also has some limitations. First, we used the MCF10A cell line, which is widely used for investigations into normal human mammary epithelial cells with the advantage of being easy to manipulate and modify, and compared to patient-derived cells, unknown genetic backgrounds in addition to BRCA1 mutations are avoided. However, the transcriptional profile and epigenetic characteristics of this cell line are distinct from those of normal human mammary cell populations and LPs (29), thus there are certain gaps between the cell model and real LP cells. Although our data show that MCF10A cells can reflect the actual functions of BRCA1 in terms of maintaining genomic stability (Fig. 1B and C) and our RNA-seq data demonstrate enrichment in the known transcriptional target genes of BRCA1, the effect of BRCA1 defects and replication stress in LPs needs to be further verified in primary human mammary cell populations. And analysis of TNBC as well as Basel like tumor data needs to be further confirmed in BRCA1-deficient breast cancer. Second, the higher replication stress levels in LPs may lead to altered expression or modifications of molecules other than ELF3, affecting the development and progression of BRCA1-associated breast cancer, which also needs to be further explored. In addition, there are likely other DEGs in our BRCA1-deficient RNA-seq that play important roles in the progression of BRCA1- associated breast cancer, and these genes also have potential research value.

Taken together, we revealed that LPs have a trait of higher replication stress, endowing them with the potential for transformation with BRCA1 deficiency. Moreover, the levels of replication stress gradually increase during tumorigenesis, indicating that replication stress is a driving factor of BRCA1-associated malignant transformation. Mechanistically, ELF3 is significantly upregulated upon replication stress and BRCA1 deficiency, which can help suppress excessive genomic instability by stabilizing the DNA replication process, empowering LPs to tolerate and to transform with BRCA1 deficiency. In cells with inherently low replicative stress, such as other non-LP mammary cells, the ELF3-associated mechanism might help cells endure the high replicative stress caused by BRCA1 deficiency without leading to cancerous changes. However, in LP cells, which naturally experience higher replicative stress, this ELF3- related mechanism may make them more susceptible to transformation into cancer cells. Notably, BRCA1 deficiency causes a tendency for cells to display characteristics of the LP transcriptional profile, and ELF3 is a core transcription factor mediating this process. Our study reveals why BRCA1 deficiency is prone to result in tumorigenesis in LPs and elucidates the key role of replication stress and ELF3 during this process, and suggests promising targets for BRCA1-associated breast cancers.

Materials and Methods

Cell culture

MCF10A cells were cultured in DMEM/F12 supplemented with 5% horse serum, 1% penicillin/streptomycin, 20 ng/mL EGF, 10 μg/mL insulin, 0.5 μg /mL hydrocortisone and 100 ng/mL cholera toxin. For MCF10A-shBRCA1-Tet-on cells, except for the replacement of horse serum with 8% Tet-on system approved certified fetal bovine serum, all other components remain the same. HEK293T and SUM149PT cells were cultured in DMEM supplemented with 10% fetal bovine serum and 1% penicillin/streptomycin. HCC1937 cells were cultured in RPMI1640 supplemented with 10% fetal bovine serum and 1% penicillin/streptomycin.

Reagents and plasmids

Reagents (antibodies and chemicals) used in this study were listed in Table S3. For the construction of Tet-on-shBRCA1-hygro plasmid, shBRCA1 sequence was subcloned into the Tet-on-pLKO-hygro vector (SinaSun). For ELF3 knockdown by lentivirus in SUM149PT, shELF3 and shCtrl sequence was subcloned into the pLKO.1-puro vector. The shRNA sequences used in this study are listed in Table S4. For ELF3 overexpression by lentivirus in MCF10A, human ELF3 was subcloned into the pCDH-CMV-MCS-EF1- puro vector, which was a kind gift from Yin lab. Myc-tagged human E2F6 was subcloned into the pDEST- N-Myc vector. PCR primers used in this study were listed in Table S5.

Transfection of siRNA and infection of shRNA

All the siRNAs were from GenePharma. For siRNA transfection, cells were seeded in 6-well plate or 35 mm2 dish at the day before transfection. 100 μL OPTI-MEM, 40 pmol siRNA and 3 μL Lipofectamine RNAiMAX reagent were mixed and added to the cells. After 48-72 h, cells were collected for subsequent experiments. The siRNA sequences used are listed in Table S6. For shRNA infection, lentivirus was generated by co-transfection of the lentiviral vector (Tet-pLKO.1-shBRCA1, pLKO.1-shCtrl and pLKO.1- shELF3) with envelope plasmids pMD2.G and psPAX2 in HEK293T cells. Lentivirus was collected after 48 h. Cells were infected with lentivirus with polybrene. After 24 h, cells were re-infected and harvested at 72 h for subsequent experiments.

Quantitative PCR (RT-qPCR)

Total RNA was extracted using Trizol reagent (Invitrogen) according to the manufacturer’s guidelines. cDNA was synthesized by reverse transcription PCR using Hifair III 1st Strand cDNA Synthesis SuperMix kit (Yeasen). qPCR was performed using SYBR Green Mix (Yeasen) on ABI StepOnePlus qPCR instrument (ThermoFisher). Each group was performed in triplicate. The relative expression mRNA levels were calculated by relative quantitative method (ΔΔCt). The primers used are listed in Table S7.

Immunofluorescence

Cells were seeded on coverslips and treated as indicated. Cells were than fixed in cold 4% paraformaldehyde for 15 min and permeabilized in 0.25% Triton X-100 solution for 5 min at room temperature. After blocking with 2% BSA for 10 min, the primary antibodies were added onto the coverslips, and then incubated for 1 h at room temperature, followed by secondary antibody incubation for 30 min. Finally, the coverslips were stained with DAPI for 5 min, and images were acquired using the ZEN software.

Immunohistochemistry

All tissue microarray chips were purchased from Shanghai Outdo Biotech Co.Ltd. IHC was performed as described previously, except that the following antibodies were used. The anti-ELF3 rabbit antibody were purchased from SIGMA. After incubation with primary antibody, the HRP-conjugated rabbit secondary antibody (ZSGB-Bio) was applied and IHC slides were observed with Olympus BX51 microscope and Olympus DP73 CCD photographic system.

Colony formation assay

MCF10A-shBRCA1-Tet-on cells were seeded in 6 mm2 plates. 1 μg/mL DOX was added for the DOX group during the whole process of cell growth to knockdown BRCA1 expression. After 14 days, the cells were washed with phosphate-buffered saline (PBS) and stained with 0.1% Coomassie brilliant blue in 10% ethanol for 30 minutes at room temperature. The stained dishes were washed with water, and the colonies were counted.

DNA fiber

The DNA fiber assay was performed as described previously with slight modifications (42). Briefly, cells transfected with siRNA were labeled by 25 μM IdU (Yuanye Bio-Technology) for 20 min and subsequently 200 μM CldU (Abcam) for 20 min. Then cells were treated with 4 mM HU for 4 h for replication fork stability assay or collected for replication fork speed and symmetry assay. Cells were resuspended in lysis buffer (200 mM Tris-HCl, pH 7.5, 50 mM EDTA and 0.5% SDS) to extract DNA. Cell lysates were then dripped onto a glass slide and DNA fibers were stretched along the slide after 3 minutes incubation. The slides were air-dried and fixed in 3:1 methanol/acetic acid solution for 10 min, followed by 2.5 M HCl treatment for 1 h, then blocked in 3% BSA PBST buffer for 20 min. DNA fibers were incubated with anti- BrdU antibody (Abcam) binding CldU and anti-BrdU antibody (BD) binding IdU at 37°C for 1.5 h. After washing with PBST, the slides were incubated with Alexa Fluor 488- and Alexa Fluor 594-conjugated secondary antibody (Invitrogen) for 45 minutes. Finally, the slides were mounted in mounting reagent (Solarbio) and DNA fibers images were acquired using ZEN software.

Comet assay

The comet assay was performed as described previously with slight modifications (43). Briefly, cells were collected by trypsin digestion and resuspended in PBS. 0.5% normal melting agarose (NMA) was added to glass slides and quickly covered with coverslips. The slides were placed at 4 °C for 10 min. The coverslips were removed carefully. The cell suspensions were mixed gently with equal volume of 1% low melting agarose (LMA) and quickly added onto the solidified NMA, quickly covered with coverslips. The slides were placed at 4 °C for 10 min. After removing the coverslips, the slides were incubated overnight in the fresh cell lysis buffer at 4 °C overnight. Then the slides were rinsed with ddH2O and electrophoresis at 4°C for 30 minutes at 20 V in neutral electrophoresis buffer. Then the slides were stained with 5 μg/ml of propidium iodide for 10 min. The comet was observed using a fluorescence microscope (Zeiss AxioCam 503 color). The pictures were analyzed using CASP software (http://casplab.com/).

Cell survival (CCK-8) assay

Cell survival assay was performed using CCK-8 kits (meilunbio, MA0218-3) according to the manufacturer’s protocols. Cells were seeded in 96-well plates. After 12 h, medium containing indicated concentration of drugs was changed and cells were incubated for 3-5 days. DMEM containing 0.1% DMSO was used as the control group. CCK-8 solution was diluted in 10% concentration and added to wells, and the cells were incubated for 45 min-1.5 h. DMEM containing 10% CCK-8 was used as the blank group. The absorbance at 450 nm was detected by a microplate reader. Cell survival was calculated according to the formula: (ODdrug – ODblank) /(ODcontrol – ODblank).

RNA sequencing

Total RNA of cells was extracted with Trizol reagent. RNA sequencing was performed by Novogene (China). The resulting data was mapped to human reference genome (hg38) and gene counts were calculated for each sample using STAR (v2.6.1) (44) with the parameter “--quantMode GeneCounts”. To identify differentially expressed genes (DEGs), genes with low expression levels (average counts per million reads < 0.05) were filtered out. DESeq2 was employed to detect DEGs with |log2 fold change| > 2 and FDR < 0.001. The raw RNA sequencing data have been deposited into GEO under the accession number GSExxxxxx. PCA analysis was performed using the “prcomp” function of stats (v3.6.1) R package.

Analysis of TCGA breast cancer and METABRIC data

TCGA (https://www.cancer.gov/tcga) breast cancer and METABRIC (19) data were downloaded using the cgdsr R package. For TCGA data, RSEM (batch normalized from Illumina HiSeq_RNASeqV2) was used as the expression levels of the genes. For METABRIC data, mRNA expression log intensity levels (Illumina Human v3 microarray) were used as the expression levels of the genes. BRCA1 associated cancers were identified as cancers with BRCA1 mutations, heterozygous loss or homozygous deletion. Pearson correlation coefficient between ELF3 expression and BRCA1 expression was calculated. Subtypes and IntClust types of breast cancers were identified according to the clinical data of TCGA and METABRIC databases.

Chip-Seq analysis of ENCODE databases

The Chip-Seq data were downloaded from the ENCODE database (https://www.encodeproject.org/) (45, 46) with the following identifiers: ENCFF858GLM, ENCFF692OYJ, ENCFF384CPN, ENCFF342GNN, ENCFF352QVM, and ENCFF437NQS. Images were generated using IGV (v2.7.2).

Drug sensitivity analysis of Cell Miner and GDSC data

Cell line drug sensitivity data and mRNA expression data were downloaded from the CellMiner (47, 48) and GDSC (49) databases. Pearson correlation coefficient between ELF3 expression and drug sensitivity was calculated for the indicated drugs.

Replication stress score and luminal progenitor score analysis

For replication stress score calculation, human mammary subgroups gene expression matrixes were obtained from the datasets of Lim et al. (3) (GSE16997). Replication stress associated pathways were selected referring to Dreyer et al. (13). and gene lists were downloaded from the GSEA website (www.gsea-msigdb.org). For each pathway, replication stress scores were calculated according to the formula below, where eg is the expression of the gene in different mammary subgroups, and ng is the number of genes in this pathway. For luminal progenitor score calculation, the LP signature gene set with average log-fold change was obtained from data of Lim et al. (3), and log CPM matrix of MCF10A-shBRCA1-Tet-on DOX 10 days vs Ctrl RNA-seq data was used. For each sample, LP scores were calculated according to the formula below, which refers to Lim et al. (3), where xg is the log-fold-change for the gene from LP gene set, and yg is the log cpm for the same gene in BRCA1 knockdown RNA-seq.

GSEA analysis and barcode plot

GSEA analysis was performed using GSEA software (v4.1.0) (50, 51) and clusterProfiler (v3.14.3) (52). For the customized gene set enrichment analysis of the transcriptional profiles of BRCA1 KD, ELF3 OE and LP, BRCA1 KD data were from MCF10A-shBRCA1-Tet-on DOX 10 days vs Ctrl RNA-Seq data and ELF3 OE data were from MCF10A OE ELF3 vs Ctrl RNA-seq data. LP gene set was obtained from 2016 cell reports, and OE ELF3 gene sets were generated using MCF10A OE ELF3 vs Ctrl RNA-seq data, where the log-fold-change cutoff was 1.3 and P value cutoff was 0.01. For the barcode plot, LP gene sets were obtained from data of Pellacani et al. (29), and log-fold-change of all genes from MCF10A- shBRCA1-Tet-on DOX 10 days vs Ctrl RNA-seq data was used for the enrichment analysis. P value was calculated using the “roast” function from limma (v3.42.2) (53).

Motif enrichment analysis

Motif matrixes of transcription factors were downloaded from the JASPAR database (54). LP gene list was obtained according to the data of Pellacani et al. (29) and LP genes are identified to significantly upregulated in LPs compared with all three other subgroups (BC, LC and SC). Promoter sequences of LP genes were extracted from the genome from 3000 bp upstream to 200 bp downstream of the TSS sites. Motif enrichment analysis was performed using SEA (Simple Enrichment Analysis) of the MEME Suite (55).

Single-cell sequencing data analysis

Single-cell RNA-seq (scRNA-seq) expression profiles of breast cancer tissue and normal mammary tissue from four BRCA1 mutation carriers, as well as normal mammary tissue from three non-carriers were generated using the 10x Genomics Chromium platform and NovaSeq 6000 platform. The detail information of sample processing, scRNA-seq data processing and quality control were described in our previous study (8). Briefly, scRNA-seq data analyses was completed utilizing the 10× Genomics Cell Ranger software suite. Principal component analysis (PCA) (twenty-five principal components) and k- means algorithm (k was set as 9.0) were used to cluster cells. We used a series of cell specific markers to identify the cell type of each cluster, as described in our previous study1. Among these, luminal progenitors cluster was identified based on the expression of KRT8/18+, CD24+, GABRP+, KIT+, and ALDH1A3+. Triple negative tumor cell population was identified based on tumor sample specific cluster, ER/PR/ HER2 expression coincident with the clinical records and PAM50 subtypes analysis. Replication stress score of each cell was calculated based on the sum expression level of genes involving in replication stress-related pathways referring to Dreyer et al. (13). In this study, we selected luminal lineage cells and tumor cells from a BRCA1-mutated triple-negative breast cancer patient to construct tumorigenesis trajectories with the R package Monocle 2 with standard settings.

Animal experiments

5×106 SUM149PT tumor cells were subcutaneously injected into the flanks of female BALB/c nude mice. Tumor growth was measured using a caliper every three days, and the tumor volume was calculated using the formula: tumor volume = length ×width2/2.

Statistics Analysis

Statistical analysis was performed using GraphPad Prism (GraphPad Software). The tests used included the paired and unpaired two-tailed Student’s t test, the Mann-Whitney U test, one-way and two-way ANOVA, and Pearson’s correlation statistical test. A P value of less than 0.05 was considered statistically significant. Details can be found in figure legends.

Study approval

The normal human breast cancer tissue and breast cancer samples used for single-cell RNA-seq in this study were collected from Peking University Cancer Hospital. This study was carried out in accordance with the ethical principles of the Declaration of Helsinki and was approved by the Research and Ethics Committee of Peking University Cancer Hospital. Written informed consent was obtained from all participants prior to participation. Animal experiments in this study were performed in accordance with the Guidelines of Peking University Animal Care and Use Committee.

Data availability

All the data supporting the findings of this study are available within the article and supplementary information file and from the corresponding author upon reasonable request. The RNA-sequencing raw data reported in this paper have been deposited in the Genome Sequence Archive in National Genomics Data Center, China National Center for Bioinformation / Beijing Institute of Genomics, Chinese Academy of Sciences (PRJCA010883) that are publicly accessible at https://ngdc.cncb.ac.cn.

Supporting information